# How this book is built {#sec-build}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

The good-practice chapter makes a case for projects, relative paths, a locked set of package versions and a build anyone else can reproduce. It would be a poor case if the book itself did not follow it, so here is the book itself, opened up. Where the arrangement is awkward, or where a choice is a judgement rather than a rule, that is said out loud.

You need none of this to work through the book. You need it to rebuild the book, to fork it for another course, or to work out why a render failed.

## The repository

```
r-intro/
├── _quarto.yml            project and book configuration
├── _common.R              sourced by every chapter, in a hidden chunk
├── _dependencies.R        never sourced, never rendered; see below
├── .Rprofile              one line, and renv wrote it
├── .gitignore             short, and one character of it is load-bearing
├── .gitattributes         keeps generated files out of diffs
├── .nojekyll              empty, and the site breaks without it
├── LICENSE                MIT
├── renv.lock              exact versions of every package
├── renv/                  activate.R, settings.json, and the library itself
├── index.qmd              the front page
├── setup.qmd …            the chapters, one file each
├── solutions.qmd          the appendices, one file each
├── data.qmd
├── build.qmd              this page
├── colophon.qmd
├── R/nus_theme.R          the ggplot2 theme and the colour palette
├── scss/nus.scss          the site styling, layered on bootswatch litera
├── scripts/check_palette.py   the palette's accessibility tests
├── assets/                favicon, and the three footer logos
├── data/sg_covid_data.csv the dataset, all 23 kB of it
├── _freeze/               cached chunk output, committed on purpose
└── docs/                  the rendered site, committed and served
```

Two of those directories are generated and both are in version control anyway, which is unusual enough that it gets its own section further down.

One file per chapter, and the chapter list in `_quarto.yml` is the only place the reading order is written down. Renaming a file means editing `_quarto.yml`, and forgetting to means the chapter silently disappears from the book while still sitting in the repository. It is the most common way to lose a page.

## `_quarto.yml`, section by section

The whole build is one YAML file. It opens with a warning to future me:

```yaml
# Verified against Quarto 1.4.554. The project: and book: schemas are closed,
# so an unrecognised key there is a hard render failure — check
# /Applications/quarto/share/schema/definitions.yml before adding one.
```

That is worth reading twice. Under `project:` and `book:`, Quarto validates keys against a fixed schema. A typo does not get ignored, and it does not warn. The render stops. The schema files shipped with your Quarto installation are the authority, and they are more reliable than a blog post about a different version.

### `project:`

```yaml
project:
  type: book
  output-dir: docs
  execute-dir: project
  resources:
    - .nojekyll
```

`type: book` is what turns a folder of `.qmd` files into a numbered book with parts, appendices, previous and next links, and a single search index.

`output-dir: docs` writes the rendered site into `docs/` rather than the default `_book/`. That one word is the entire publishing strategy, and there is a section on it below.

`execute-dir: project` sets R's working directory to the project root for every chunk in every chapter. Without it, each chapter would execute in its own directory, which is fine here because every file sits at the top level, but stops being fine the moment anyone puts a chapter in a subfolder. Setting it means `here::here("data", "sg_covid_data.csv")` resolves identically from every chapter, including this one.

`resources:` lists files to copy into `docs/` on every project render. `.nojekyll` is an empty file whose entire job is explained in the publishing section.

### `book:`

Metadata first: title, subtitle, author, `date: today` with an explicit `date-format`, and a description that becomes the page's meta description and the social preview text.

`site-url` and `repo-url` together are what make the "Edit this page", "Report an issue" and "View source" links work. `repo-actions: [edit, issue, source]` turns all three on. They appear in the right-hand margin of every page, and they are the cheapest error-reporting mechanism a teaching resource can have. A student who spots a wrong number is two clicks from telling me, and one of those clicks lands them in a text editor with the offending line in front of them.

`repo-branch: main` matters because Quarto builds the edit URL from it. Point it at a branch that does not exist and every edit link 404s, on every page, silently.

Then the reading conveniences: `search: true`, `reader-mode: true` (the button that collapses both sidebars), `page-navigation: true` and `back-to-top-navigation: true`.

The sidebar is docked rather than floating, collapsed to one level so the parts are visible but the sections are not, and carries two tools: a link to the source on GitHub, and a direct download link for the CSV. The download link points at `raw.githubusercontent.com`, not at the repository page, so it downloads the file instead of showing a preview of it.

### `chapters:` and `appendices:`

```yaml
  chapters:
    - index.qmd
    - part: "Before you write any code"
      chapters:
        - setup.qmd
        - good-practice.qmd
```

The parts are headings in the sidebar and nothing else; they carry no files of their own. `index.qmd` sits outside any part because a book's front page is not a chapter.

`appendices:` is a separate list, and Quarto numbers it separately: chapters get 1, 2, 3, appendices get A, B, C, D. The `crossref: appendix-title: "Appendix"` setting at the bottom of the file is what makes a reference to one of them read "Appendix B" rather than just "B".

### `page-footer:`

The footer carries the three institutional logos as a raw HTML block, plus the licence on the left and the Quarto credit and issue link on the right. The logos are `<img>` tags pointing at `assets/logos/`, styled by a `.footer-logos` rule in `scss/nus.scss`. They are raster PNGs rather than SVGs because that is the artwork as supplied, and redrawing an institution's mark is not a thing to do casually.

### `format:`

```yaml
format:
  html:
    theme: [litera, scss/nus.scss]
```

Bootswatch `litera` underneath, the project's own SCSS on top. `litera` was chosen for a reason that has nothing to do with taste: it is one of the few bundled themes that does not fetch a web font from `fonts.googleapis.com`. The book renders the same on a plane as it does on a fast connection. For a course whose first session is often held on university wifi, that is worth more than a nicer typeface.

The rest of the HTML block, in the order it appears:

`toc: true` with `toc-depth: 3` and a custom `toc-title`. `number-sections: true`, so sections can be referenced by number. `anchor-sections: true` gives every heading a link target, which is what lets me point a student at one paragraph rather than a chapter. `smooth-scroll: true` is cosmetic.

`code-copy: true` puts a copy button on every code block. `code-overflow: wrap` wraps long lines instead of adding a horizontal scrollbar, because a line that scrolls off the right edge on a phone is a line nobody reads.

`code-link: true` is the good one. It turns every function call in every code block into a link to that function's documentation, so `filter()` in a chunk is clickable and lands on the dplyr reference page. It needs `downlit` and `xml2` installed. Neither is called by name anywhere in the book, so both are in the lockfile only because `_dependencies.R` names them, which is a trap the renv section comes back to.

`code-annotations: hover` is how the numbered callouts in code blocks behave.

`df-print: default` is a deliberate refusal. Set it to `kable` and every data frame in the book renders as a tidy HTML table with proper borders. It would look better. It would also look nothing like what the student sees in their own console, and a beginner comparing the book to their screen would conclude they had done something wrong. Console output stays console output.

Figures: `fig-width: 7`, `fig-height: 4.2`, `fig-align: center`, `fig-format: svg`. SVG because plot text stays sharp at any zoom and the files are usually smaller than PNG for this kind of chart. `tbl-cap-location: bottom` follows the usual convention of captions below tables and above figures.

### `execute:` and why chapters must not fight it

```yaml
execute:
  freeze: auto
  echo: true
  warning: true
  message: true
```

Warnings and messages are on, globally and on purpose. `read_csv()`'s column specification, `geom_histogram()`'s note about picking 30 bins, and `summarise()`'s `.groups` notice are all messages, and the book teaches students to read all three. A chapter that sets `message: false` to make its output tidy is deleting the lesson.

Because these are set globally, chapters should not set them again. Repeating `#| echo: true` on a chunk is harmless but noisy; repeating it inconsistently is how a chapter ends up with three different behaviours for no reason anybody can reconstruct later. The same goes for `comment`, `collapse`, and the figure dimensions, all of which are set once in the `knitr:` block at the bottom:

```yaml
knitr:
  opts_chunk:
    comment: "#>"
    collapse: true
    dev: "svg"
```

`comment: "#>"` prefixes every line of output. It means you can select a whole block, code and output together, paste it into your console, and have it run: the output lines are comments. `collapse: true` puts code and its output in one block rather than two.

Two per-chunk options are used, and only these two. `#| eval: false` for anything that must not run during a render, which means `install.packages()`, `renv::` calls, `setwd()` demonstrations, `View()`. `#| error: true` for chunks that are supposed to fail, where the error message is the point.

## `_common.R`

Every chapter opens with the same four lines:

````
```{{r}}
#| label: setup
#| include: false
source("_common.R")
```
````

`include: false` hides the chunk and its output. The reader never sees it.

The repetition is not laziness. Quarto's freeze machinery gives each chapter its own R session, which is what makes a single chapter re-renderable without re-running the whole book. A variable created in chapter 4 does not exist in chapter 5. So anything more than one chapter needs has to be defined in a file that every chapter sources, and `_common.R` is that file. It attaches the packages, sets the ggplot2 theme, defines `read_covid()` and `covid_daily()`, and runs the guard described in the next section.

The leading underscore is a Quarto convention: files and directories starting with `_` are not treated as book content and are not copied to the output.

## renv, and the one that will get you

`renv` records the exact version of every package the book was built with. It does this in `renv.lock`, a JSON file with one entry per package:

```json
    "here": {
      "Package": "here",
      "Version": "1.0.2",
      "Source": "Repository",
      ...
      "Repository": "CRAN"
    },
```

(Trimmed. Each real entry also carries the package's title, licence, dependencies and maintainer.)

The lockfile is a manifest, not a copy of the packages. It is a few hundred kB of text that says which versions to go and get.

```{r}
#| label: lockfile
lock <- renv::lockfile_read()
lock$R$Version
length(lock$Packages)
```

`r length(renv::lockfile_read()$Packages)` packages, for a book whose `_common.R` calls `library()` four times. Almost all of them are dependencies of dependencies, which is the honest reason to use a lockfile at all: nobody can hold that graph in their head, and "install the tidyverse" resolves to a different set of versions every month.

The three commands you need:

`renv::restore()` reads the lockfile and installs exactly those versions into `renv/library`. This is the first thing to run in a fresh clone. It takes a few minutes once.

`renv::status()` compares what is installed against what is recorded, and tells you which way they disagree.

`renv::snapshot()` writes the current state back into the lockfile. Run it after adding a package, and commit the resulting diff along with the change that needed it.

The library itself is not in version control. renv writes and maintains `renv/.gitignore` for you:

```
library/
local/
cellar/
lock/
python/
sandbox/
staging/
```

`renv.lock`, `renv/activate.R` and `renv/settings.json` are tracked. Those three plus the ignore rules are what makes a clean checkout reproducible without committing several hundred megabytes of compiled code.

`.Rprofile` is one line, and renv wrote it:

```r
source("renv/activate.R")
```

R runs `.Rprofile` from the working directory at startup. That line activates the project library, and it is the whole mechanism.

### `_dependencies.R`

renv finds dependencies by scanning source files for `library()` calls. Its scanner automatically adds `rmarkdown` when it sees `.Rmd` files, but not for `.qmd`. This book has no `.Rmd` files, so without intervention `renv::snapshot()` leaves `knitr` and `rmarkdown` out of the lockfile entirely, and a clean checkout fails to render with an error that points nowhere near the cause.

`_dependencies.R` is a file that exists only to be scanned. It is never sourced and never rendered. It lists the packages the chapters use plus the ones the toolchain needs but nothing calls by name: `knitr`, `rmarkdown`, `downlit`, `xml2`, `sessioninfo`. If you add a package to a chapter, add it there too, or it will work on your machine and nowhere else.

### The gotcha

This is the one that cost me an evening.

Quarto decides which renv to activate by reading `.Rprofile` **relative to the directory you invoked it from**, not relative to the project you asked it to render. So this is fine:

```bash
cd ~/r-intro
quarto render
```

and this is not:

```bash
cd ~/sph5421-course      # another R project, with its own .Rprofile and its own renv
quarto render ../r-intro
```

The second one starts R in `~/sph5421-course`, reads that project's `.Rprofile`, activates that project's library, and then renders this book against whatever versions happen to be sitting there. `execute-dir: project` moves the working directory afterwards, so `here()` and every relative path still work, and the render completes.

That is what makes it dangerous. The versions in a neighbouring project are usually close enough that most of the book builds, and you find out three chapters later when a `dplyr` deprecation warning appears on a page that never had one.

So `_common.R` checks, before it does anything else:

```r
local({
  proj <- getwd()
  while (!file.exists(file.path(proj, "_quarto.yml")) && dirname(proj) != proj) {
    proj <- dirname(proj)
  }
  expected <- normalizePath(file.path(proj, "renv", "library"), mustWork = FALSE)
  active   <- normalizePath(.libPaths()[1], mustWork = FALSE)

  if (!startsWith(active, expected)) {
    stop(
      "Wrong package library.\n",
      "  active:   ", active, "\n",
      "  expected: ", expected, "/...\n",
      "Run `quarto render` from the project root (", proj, ").",
      call. = FALSE
    )
  }
})
```

It walks up from the working directory until it finds `_quarto.yml`, works out where that project's library ought to be, and compares it against the library R is actually using. When they disagree the render stops on the first chapter with:

```
Error: Wrong package library.
  active:   /Users/you/sph5421-course/renv/library/macos/R-4.6/aarch64-apple-darwin25.4.0
  expected: /Users/you/r-intro/renv/library/...
Run `quarto render` from the project root (/Users/you/r-intro).
```

The two paths will be yours. The fix is always the same: `cd` into the project root and render from there.

Here is the library this page was built against:

```{r}
#| label: libpath
.libPaths()[1]
```

If that path does not contain `r-intro/renv`, the guard has a bug, which would also be worth an issue.

## Freeze

`freeze: auto` means a chapter's R code is executed when the chapter's `.qmd` changes, and not otherwise. The output of every chunk is cached under `_freeze/`, and `_freeze/` is committed.

The payoff is that anyone can clone the repository and rebuild the site without R, without renv, and without the several minutes of computation. It also means a one-word fix to chapter 9 does not re-execute chapters 1 through 8, which turns a two-minute render into a five-second one and is the difference between previewing your changes and not bothering.

The cost is a trap. **Freeze watches the `.qmd` file, not the things it depends on.** Edit `_common.R`, change the colour palette in `R/nus_theme.R`, or replace the CSV, and nothing re-runs. The book will render happily and every page will still show the old output. There is no warning, because from Quarto's point of view nothing changed.

There is no command-line flag to force it. When you have changed something shared, delete the cache:

```bash
rm -rf _freeze/
quarto render
```

or delete one chapter's directory under `_freeze/` if you know that only one is affected. Doing this after any change to `_common.R` is a habit worth building, and the reason `_freeze/` is listed in `.gitattributes` as generated is that the resulting diff is enormous and nobody should read it.

## Publishing

The site is rendered locally into `docs/`, and `docs/` is committed. GitHub Pages is configured to deploy from a branch: `main`, folder `/docs`. Push, wait a minute, done. There is no CI workflow, no `gh-pages` branch, and no build step on anybody's server.

That is not the fashionable choice. The alternative is `quarto publish gh-pages`, or a GitHub Action that installs R, restores renv, renders, and pushes. Both are better for a project with several contributors. This one is better here for two reasons. The rendered page and the source that produced it land in the same commit, so `git log` on a chapter shows you exactly what the site looked like at every point. And there is no CI to break: a course resource that stops publishing in the middle of teaching week because an action's Ubuntu image changed is a worse failure than a slightly larger repository.

`.nojekyll` is the other half. GitHub Pages runs Jekyll over your site by default, and Jekyll ignores every file and directory whose name begins with an underscore. Quarto's HTML output is full of them. An empty `.nojekyll` file at the root turns Jekyll off entirely, and listing it under `project: resources:` is what copies it into `docs/` on every render, including after `quarto render` cleans the output directory. Miss it and the site deploys with no CSS, which looks exactly like a broken theme and is not.

### The leading slash in `.gitignore`

```
# The leading slash is load-bearing. An unanchored `*_files/` also matches
# docs/<chapter>_files/, which is where every rendered figure lives — the site
# would look right locally and ship with every image broken.
/*_files/
```

Knitr writes a chapter's figures into `<chapter>_files/` next to the source, and those working directories are rubbish that should not be in git. The obvious rule is `*_files/`.

The obvious rule is wrong. A pattern with no slash at the start matches at **any depth**, so `*_files/` also matches `docs/visualise_files/`, `docs/groups_files/`, and every other directory in the rendered site that holds the actual published images. Git would ignore all of them. You would render, see a perfect site on `localhost`, commit, push, and get a live site where every plot is a broken-image icon. Nothing in your local build would ever show the problem.

The leading slash anchors the pattern to the repository root, so it matches `./visualise_files/` and stops there. One character.

`.gitattributes` finishes the job, marking both generated directories so they stay out of GitHub's language statistics and out of diffs:

```
docs/** linguist-generated=true
docs/** -diff
_freeze/** linguist-generated=true
_freeze/** -diff
```

Without that, GitHub reads the repository as mostly HTML and JavaScript rather than R, and a one-line prose fix arrives as a pull request with thousands of changed lines.

## Forking it

The licence says you can, so here is how.

```bash
git clone https://github.com/CERM-NUS/r-intro.git my-course
cd my-course
```

Open `r-intro.Rproj` in RStudio. The `.Rproj` file keeps its old name after a clone, so rename it now if the old one will confuse you later. Then, in the R console:

```r
renv::restore()
```

Answer yes. Go and make coffee. When it finishes, from a terminal **in that directory**:

```bash
quarto preview
```

That builds the book and opens it in a browser with live reload: save a `.qmd` and the page updates. It is the right mode for writing. `quarto render` is the right mode for publishing.

Four things to change before you publish your version:

`_quarto.yml` needs a new `site-url`, `repo-url` and probably `repo-branch`, or every edit link on your site will point at mine.

The logos in `assets/logos/` have to go, or be replaced with your own. They are the trade marks of NUS, the Saw Swee Hock School of Public Health, CERM and the ML & GHN network, and the MIT licence on the text and code does not cover them. Remove the `page-footer` HTML block along with the files.

`LICENSE` should keep the original copyright line if you keep any of the original text, and add yours.

If you swap the dataset, `_common.R` is the only place that reads it, but `data.qmd` documents the columns of *this* file and will be wrong the moment you change it. Rewrite that appendix or delete it. An out-of-date data appendix is worse than none.

Then render, commit `docs/` and `_freeze/` along with your sources, push, and switch on Pages for `/docs`.
