# A monthly situation report {#sec-capstone}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

```{r}
#| label: data
#| include: false
covid <- covid_daily()
```

Up to now every chapter has handed you a question with the shape of the answer already showing:
filter this, group by that, plot the result. Requests in practice are less specified. They arrive
from someone who wants to know what happened, by a deadline, in a form they can put in front of a
committee.

The request in this chapter is of that kind: pick a month of Singapore's 2020 and write the
report. April is worked through first, to set out what the finished report looks like, and August
is left to you.

## The brief

The report covers one month and occupies one page. That page carries a table that gives the
reader the month in numbers, two figures that show something the table cannot, and three
sentences of interpretation. Three sentences, not three paragraphs: the limit is deliberate,
because choosing which three findings matter takes more work than writing several paragraphs of
hedged description.

Assume the reader is a public health officer who knows Singapore and does not know R. They will
not read your code. They will look at the figures first, the table second, and your sentences
only if the figures made them curious.

## What a good report contains

There are six things I look for, and they are all things you already know how to do.

1. **The month is stated, and so is the denominator.** How many days does the month have, and
   how many of them carry data? Every month in this file is complete, but state that rather than
   leaving the reader to assume it.
2. **A count and a shape.** A total tells you the size of the month. A median, a maximum and
   the date of the maximum tell you whether that total accumulated steadily or came from a small
   number of large days. Report both.
3. **A table at a unit the reader can take in.** Days are too fine for a monthly table and the
   month itself is too coarse. Weeks are the workable grain.
4. **Two figures that answer different questions.** Two views of the case curve is one figure
   drawn twice. Pair the epidemic curve with something else in the file: mobility, the
   stringency index, the severity mix.
5. **Every figure has a title that states a finding.** "Daily cases, April 2020" is a label.
   "Cases rose for 13 days after the circuit breaker began" is a finding. A title that makes a
   claim has to be checked against the data before it can stand.
6. **At least one caveat that comes from looking at the data, not from a template.** "Data may
   be incomplete" is a template. "The 5 August count is four times the next highest August day
   and needs an explanation before it goes in a mean" is analysis.

## April 2020, worked

### A script that stands on its own

Your report script must be readable from the top without any hidden state. That means it reads
the file and builds its own columns, in view, before it uses them. This is exactly the pipeline
from the conditionals chapter, written out once so nothing in the report depends on something
you cannot see.

```{r}
#| label: build
covid <- read_covid() %>%
  arrange(date) %>%
  mutate(
    daily_cases  = confirmed - lag(confirmed, default = 0),
    daily_deaths = deaths - lag(deaths, default = 0),
    severity = case_when(
      daily_cases <  100 ~ "Low",
      daily_cases <= 300 ~ "Medium",
      daily_cases >  300 ~ "High"
    ),
    severity = factor(severity, levels = c("Low", "Medium", "High")),
    month = floor_date(date, "month")
  )
```

If you are writing this in a script of your own rather than inside the book's project, swap
`read_covid()` for `read_csv(here("data", "sg_covid_data.csv"))`. That helper exists so that the
chapters do not repeat a file path in every chunk, and nothing else in the pipeline depends on it.

Then pick the month once, at the top, in a named object. When you move to August you change one
line rather than hunting for `"2020-04"` scattered through twenty chunks.

```{r}
#| label: pick-month
report_month <- as.Date("2020-04-01")

april <- covid %>%
  filter(month == report_month)

nrow(april)
```

Thirty rows for a thirty-day month. The check is one line, and it catches a mis-specified filter
before anything is computed from it.

### The headline numbers

::: {.callout-important}
## Predict before you run this

April 2020 has one day far larger than the rest. Write down the date you think it fell on before
you read the next output. Most people put it in the first week, on the reasoning that the circuit
breaker started on 7 April and things improved from there.
:::

```{r}
#| label: headline
april %>%
  summarise(
    days         = n(),
    cases        = sum(daily_cases),
    mean_daily   = mean(daily_cases),
    median_daily = median(daily_cases),
    peak         = max(daily_cases),
    peak_date    = date[which.max(daily_cases)],
    deaths       = sum(daily_deaths)
  )
```

The peak is 20 April, thirteen days after the circuit breaker began. Reported cases continued to
rise for two weeks after Singapore closed workplaces and required people to stay at home.

That is not a paradox and it is worth being precise about why. Reported cases are the output of
a testing programme, not a direct measurement of infection. Through April, testing in the
migrant-worker dormitories was expanding fast, and the dormitory outbreak dominated the national
count from April to August. A rising reported curve during that period is partly transmission
and partly the swab rate catching up with infections that had already happened. The two cannot be
separated from this file, and the report should say so.

Two small notes on that output. `median_daily` prints as `488.` with a trailing full stop, which
is the tibble telling you the value is not a whole number. It is 487.5, the average of the two
middle days. And `deaths` is 12 here, which is reliable for April because the `deaths` column
has no missing values from 21 March onwards. Earlier in the year it does, 59 leading `NA`s, and
lag-differencing across them recovers only 27 of the year's 29 deaths. If your chosen month is
January, February or March, state that in the report.

```{r}
#| label: severity-mix
april %>%
  count(severity)
```

Eighteen of April's thirty days were over 300 cases. Across the whole year only 77 days were,
so April holds nearly a quarter of the year's highest-count days in a single month.

### The table

Weeks are the right grain for a monthly table. `floor_date(date, "week", week_start = 1)` puts
each date on its Monday. I set `week_start` explicitly because lubridate's default is Sunday and
many readers will expect Monday. Setting it explicitly makes the assumption visible at the cost
of one argument.

```{r}
#| label: weekly-table
april_weeks <- april %>%
  mutate(week = floor_date(date, "week", week_start = 1)) %>%
  group_by(week) %>%
  summarise(
    days       = n(),
    cases      = sum(daily_cases),
    mean_daily = round(mean(daily_cases)),
    peak       = max(daily_cases),
    retail     = round(mean(retail_and_recreation_percent_change_from_baseline)),
    stringency = round(mean(stringency_index), 1)
  )

april_weeks
```

Look at the `days` column before you look at anything else. The first row covers five days, not
seven, because April 2020 started on a Wednesday and the week beginning 30 March mostly belongs
to March. The last row covers four. Any comparison of `cases` across those rows is comparing
different amounts of time, which is why `mean_daily` is in the table and why `days` is in it too.
Leave `days` out and a reader will read 383 against 1,223 as a threefold rise when part of the
difference is two extra days.

The rest of the table tells a fairly clean story. Weekly totals climb from 383 to 7,036 and then
fall back. Mean retail mobility drops from 23% below baseline in the first partial week to 67%
below in each of the last two. Mean stringency goes from 41.7 to 71 in one step and sits in the
seventies for the rest of the month.

### The first figure: the epidemic curve

```{r}
#| label: fig-april-cases
#| fig-cap: "Daily reported cases in Singapore, April 2020, with the start of the circuit breaker marked."
cb_start <- as.Date("2020-04-07")

ggplot(april, aes(date, daily_cases)) +
  geom_col(fill = nus_colours[["blue"]]) +
  geom_vline(xintercept = cb_start, colour = nus_colours[["crimson"]],
             linetype = "dashed") +
  annotate("text", x = cb_start + 1, y = 1350, hjust = 0, size = 3.2,
           colour = nus_colours[["crimson"]], label = "Circuit breaker begins") +
  scale_x_date(date_breaks = "1 week", date_labels = "%d %b") +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Cases rose for 13 days after the circuit breaker began",
    subtitle = "Singapore, new COVID-19 cases reported each day, April 2020",
    x = NULL, y = "New cases reported",
    caption = "Source: COVID-19 Data Hub. Reported cases, not infections."
  )
```

Columns rather than a line, because these are counts of things that happened on discrete days
and a line implies you could read a value off it halfway between two dates. The dashed rule and
the label do the work that a paragraph of text would otherwise have to do. The caption carries
the source and the one warning the reader most needs.

The date breaks are weekly and the labels are `"%d %b"`, so the axis reads 06 Apr, 13 Apr, 20
Apr. Left to itself ggplot2 would have chosen something reasonable but not necessarily aligned
with the weeks in your table, and a reader who is trying to match figure to table will notice.

### The second figure: mobility

The case curve records what was reported. The mobility columns record where people went, which is
a different question and the reason this file supports more than a case series on its own.

Two series on one set of axes means one row per series per date, so the two columns have to
become one column of values plus one column of labels. `pivot_longer()` does that. It is from
tidyr, which came in with the tidyverse, and it is the only function in this chapter you have not
already met.

```{r}
#| label: mobility-long
april_mobility <- april %>%
  select(
    date,
    `Retail and recreation` = retail_and_recreation_percent_change_from_baseline,
    Workplaces              = workplaces_percent_change_from_baseline
  ) %>%
  pivot_longer(-date, names_to = "series", values_to = "change")

april_mobility
```

Renaming inside `select()` is what gets you a readable legend without a second call. Backticks
are needed around `Retail and recreation` because it contains spaces; R will accept a name like
that as long as you quote it that way, and it is worth doing here because that string goes
straight onto the plot.

```{r}
#| label: fig-april-mobility
#| fig-cap: "Retail and workplace mobility in April 2020, as percent change from Google's pre-pandemic baseline."
ggplot(april_mobility, aes(date, change, colour = series)) +
  geom_hline(yintercept = 0, colour = nus_colours[["muted"]], linewidth = 0.3) +
  geom_line() +
  geom_vline(xintercept = cb_start, colour = nus_colours[["crimson"]],
             linetype = "dashed") +
  scale_x_date(date_breaks = "1 week", date_labels = "%d %b") +
  labs(
    title = "Movement fell the day the rules changed, and stayed down",
    subtitle = "Percent change from Google's 3 January to 6 February 2020 baseline",
    x = NULL, y = "Change from baseline (%)", colour = NULL,
    caption = "Weekends are not comparable with weekdays: the baseline is a median by day of week."
  )
```

The step at 7 April is what the figure is for. Retail goes from 15% below baseline on 6
April to 51% below on 7 April; workplaces go from 24% below to 52% below on the same day.
Behaviour changed within a day, while the case curve took a fortnight to turn.

Two features of that figure are not what they appear to be. The deep spike down to 83%
below baseline is 10 April, which was Good Friday, a public holiday. The eight deepest
workplace-mobility days in the entire year are all public holidays, so any dip you see in this
series should be checked against a calendar before it is called a policy effect.

The regular bumps upwards are weekends, and they run the opposite way to intuition. Google's
baseline is a median computed separately for each day of the week, so a Sunday is compared with
pre-pandemic Sundays. Workplace activity on a normal Sunday was already low, which leaves less
room to fall.

```{r}
#| label: weekday-check
april %>%
  filter(date >= cb_start) %>%
  mutate(weekday = wday(date, label = TRUE)) %>%
  group_by(weekday) %>%
  summarise(
    days       = n(),
    workplaces = mean(workplaces_percent_change_from_baseline)
  )
```

Weekdays sit between 63 and 72 points below baseline. Saturday and Sunday sit around 50. Workplace
activity did not return at the weekend; the baseline it is compared against is lower.

If you want the two figures stacked as a single image for a one-page PDF, patchwork will do it
with `p_cases / p_mobility` once you have assigned each plot to a name. I have kept them separate
here so that each keeps its full height on screen.

### The three sentences

> April carried 15,243 reported cases, 26% of Singapore's 2020 total, with 18 of its 30 days
> above 300 cases and 12 deaths; only May, at 18,715, was heavier.
>
> Reported cases kept climbing for 13 days after the circuit breaker took effect on 7 April,
> peaking at 1,426 on 20 April, a pattern that reflects the pace of testing in the
> migrant-worker dormitories as much as the timing of infection.
>
> Movement changed the day the rules did, with retail activity falling from 15% below baseline
> on 6 April to 51% below on 7 April and staying within a few points of its floor for the rest
> of the month, so the behavioural response was immediate even though the case curve was not.

The three sentences do three different jobs. The first sizes the month against the year. The
second addresses the point a reader would otherwise misread. The third brings in the second data
source and sets the behavioural response against the case curve.

Two things are absent. There is no recommendation, because nothing in this file supports one, and
no claim that the circuit breaker did or did not work, because a single month of a single country
with no counterfactual cannot establish either. The purpose of a situation report is description
rather than causal inference.

## What this analysis does not tell you

Write a version of this section for your own month, and make it specific to that month.

**Raw counts ignore population.** Fifteen thousand cases is a number of people, not a rate. To
turn it into a rate you need a denominator, and the appropriate denominator here is not obvious.
Roughly 323,000 work-permit holders lived in dormitories of all types in 2020, of whom about
200,000 were in the 43 purpose-built dormitories. A case rate computed over the national
population and a case rate computed over the dormitory-resident population are different
quantities, and in April 2020 they were not close. This file gives you neither
denominator.

**Raw counts ignore testing volume.** A reported case requires a test, a positive result and a
report. Change the number of tests and the case count changes without a single extra infection.
There is no test-count column in this file, so you cannot compute test positivity, which is the
standard way of asking whether a rise in cases reflects a rise in transmission or a rise in
testing. A rising curve supports the statement that reported cases rose. It does not support the
statement that infections rose.

**One national count is averaging two epidemics.** From April to August 2020 the dormitory
outbreak dominated Singapore's case counts, and it had different dynamics from transmission in
the wider community: different living density, different age structure, different testing
regime, different intervention. MOH's own daily updates reported those streams separately at the
time. This file carries a single national total, so every mean you compute over those months is
a weighted average of two processes, with weights you cannot see and that change from week to
week. That is a limitation of the file, not of the epidemiology, and you should say which one
you are reporting.

5 August 2020 is the clearest example. That day records 908 cases, against an August median of
91 and a next-highest August day of 313. Nine hundred and three of those 908 were dormitory
residents, and 581 were linked to a single cluster at Acacia Lodge. MOH attributed the day's
numbers to clearance testing in dormitories picking up past infections. A mean that includes that
day without comment describes a change in reporting as though it were a change in incidence.

One further point, flagged as my own reading rather than something you can verify from the
data. Google's mobility series are built from location history on the phones of users who have
that setting switched on, and Google does not publish these series broken down by nationality or
work-permit status. Whether the movement of people confined to dormitories under quarantine
appears in these curves at all is not a question this file can answer. I would not rely on the
mobility series to describe what happened inside the dormitories in April. R will compute a mean
of any column it is given; whether the column supports the question is a separate judgement.

## Your turn

::: {#exr-sitrep}
## Write the August 2020 situation report

Produce the one-page report for August 2020: a table, two figures, three sentences of
interpretation, and a short note on what your analysis does not tell you.

August is a harder month than April, though not because the code is harder; the code is the same.
August is harder because the month contains a decision that April does not, and the report has to
address that decision explicitly.

You are done when someone who has never seen the data could read your page and come away with a
correct impression of what August 2020 was like.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

Compute the maximum, the median and the date of the maximum before you compute anything else,
and put those three numbers next to each other. If a single day is several times the next
highest day, the mean is describing that day rather than the month, and your report has to
state explicitly what it does about that.
:::

[Worked solution](solutions.qmd#sec-sol-sitrep)

If you want a second month after that, September 2020 is the quiet counterpart: 30 days, every
one of them under 100 cases, a mean of 31.8. Reporting on a month in which little changed is a
separate skill, and one that is required regularly.
