# Objects, vectors and functions {#sec-foundations}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

Type `2 + 2` at the console and R prints `4`. Type it again and R prints `4` again. Nothing has been kept, and nothing can be built on. Before you can do anything with 345 days of case counts, R has to hold on to things and give them names, and you have to know what kind of thing it is holding.

That last part is where beginners lose whole evenings. R will happily let you put a word into a column of numbers, and it will not shout. It will convert the numbers into words instead, and then your average comes back as `NA` and you have no idea why. This chapter is mostly about giving you the vocabulary to diagnose that in ten seconds rather than an hour.

The Singapore file is used throughout, so load it now. `read_covid()` is a small helper that ships with this book; @sec-import pulls it apart and shows you what it does. For the moment it hands you the data.

```{r}
#| label: load
covid <- read_covid()
```

## Making R keep something {#sec-assignment}

Assignment gives a value a name. R's assignment operator is a left-pointing arrow, made of a `<` and a `-`:

```{r}
#| label: first-assign
first_case <- as.Date("2020-01-23")
first_case
```

Two things happened. R evaluated the right-hand side, and it bound the result to the name `first_case`. Assignment produces no output, which surprises people; if you want to see the value, type the name on its own line, as above.

That date is the first day this file records a confirmed case. Day one of the file, 22 January, has `confirmed` equal to zero.

In RStudio, `Alt` and minus (`Option` and minus on a Mac) types `<-` for you. Learn it early; you will type this operator thousands of times.

### Why an arrow and not an equals sign

R inherited `<-` from S, and S inherited it from APL, whose keyboards had a single key that produced a real `←` character. When S moved to keyboards without that key, the arrow was spelled out as two characters and stuck.

`=` also works for assignment at the top level. This book does not use it, for one reason: inside a function call, `=` already means something else. In `round(x, digits = 2)` the `digits = 2` is not creating an object called `digits`. It is telling `round()` which argument you are filling. Keeping `<-` for "make an object" and `=` for "fill an argument" means you can read a line of code and know which is happening without thinking about it.

### Naming things

Names may contain letters, numbers, dots and underscores. They cannot start with a number, and they cannot contain spaces or hyphens. `daily_cases` is fine, `daily cases` is a syntax error, and `daily-cases` is read as `daily` minus `cases`.

R is case sensitive. `Confirmed` and `confirmed` are different objects, and mixing them up produces the single most common error message in R:

```{r}
#| label: not-found
#| error: true
covd
```

When you see `object 'x' not found`, R is telling you it looked everywhere it knows about and there is no such name. Nine times out of ten it is a typo or a chunk you never ran.

This book uses `snake_case`: lower case, words joined by underscores. The tidyverse does the same, so your code will look like the code around it.

Two names to avoid. `TRUE` and `FALSE` are reserved and cannot be reassigned, but their abbreviations `T` and `F` are ordinary variables. Someone who writes `T <- 0` early in a long script has quietly redefined truth for everything that follows. Write `TRUE` and `FALSE` in full. Similarly, avoid naming your own objects `c`, `data`, `mean` or `df`; R will usually cope, but you will confuse yourself.

## R has no scalars {#sec-vectors}

Here is the fact that makes the rest of R make sense. There is no such thing in R as a single number. What looks like one is a vector of length one.

```{r}
#| label: length-one
total_deaths <- 29
length(total_deaths)
length("Singapore")
```

The vector is the atom. `c()` (for *combine*) glues values into a longer one:

```{r}
#| label: combine
week_one <- c(0, 1, 3, 3, 4, 5, 7)
week_one
length(week_one)
```

Every column of the Singapore file is a vector of length 345. Pull one out with `$`:

```{r}
#| label: dollar
confirmed <- covid$confirmed
length(confirmed)
head(confirmed, 10)
```

Because vectors are the atom, arithmetic works on whole vectors at once. R does not need you to write a loop over 345 days.

```{r}
#| label: vectorised
head(confirmed * 2, 10)
```

This is called vectorisation, and it is the reason R code for data analysis is short. Later, when `mutate()` computes a new column for every row of a data frame in one line, it will feel unremarkable, because you already know that arithmetic on a vector means arithmetic on all of it.

Here is the same idea doing something useful. The `confirmed` column is cumulative, a running total. The number of *new* cases on a given day is that day's total minus the previous day's. Shift the vector along by one, stick a zero on the front so the lengths still match, and subtract:

```{r}
#| label: daily-by-hand
previous_day <- c(0, confirmed[-length(confirmed)])
new_cases    <- confirmed - previous_day
head(new_cases, 10)
max(new_cases)
covid$date[which.max(new_cases)]
```

1,426 new cases on 20 April 2020, the peak of the dormitory outbreak. Three lines, no loop, and the answer is right: the new cases add back up to the final total.

```{r}
#| label: daily-check
sum(new_cases) == max(confirmed)
```

The zero on the front is doing real work there, and it is not a fudge. The file starts on 22 January with `confirmed` equal to zero, so 22 January really did have zero new cases. Writing that day as missing instead would drop a genuine day from every count you make afterwards. `dplyr::lag()` does this shifting properly and you will meet it soon; the version above is the same idea with the machinery visible.

### Subsetting

Square brackets pick elements out by position, counting from 1 rather than 0.

```{r}
#| label: subset
confirmed[1]
confirmed[1:5]
confirmed[length(confirmed)]
```

A negative index means *everything except*, which is what `confirmed[-length(confirmed)]` did above: all of it bar the last day.

You can also subset with a logical vector, which is where this starts to pay off:

```{r}
#| label: logical-subset
covid$date[new_cases > 1000]
```

Four days in 2020 with more than a thousand new cases, and they run consecutively from 20 April. `filter()` will do this more readably later, but underneath it is this.

## Four types, and one absence {#sec-types}

Every vector has a type, and every element of a vector has the *same* type. Four types matter for data work:

```{r}
#| label: typeof
typeof(TRUE)
typeof(2L)
typeof(29)
typeof("Singapore")
```

Logical, integer, double, character. (There are two more, `complex` and `raw`. You can ignore them for the whole of your career unless you go looking.)

Integer and double are both numbers, and R makes the distinction awkwardly. A number you type is a double unless you stick an `L` on the end. A colon range is an integer.

```{r}
#| label: int-dbl
typeof(1:5)
typeof(5 / 2)
identical(1, 1L)
1 == 1L
```

`identical()` says they are different objects; `==` says they are the same value. Both are correct, and in practice you almost never need to care. Where it bites is that doubles are stored in binary floating point, which cannot represent most decimal fractions exactly:

```{r}
#| label: float
0.1 + 0.2 == 0.3
0.1 + 0.2 - 0.3
```

That is not R being broken. Every language using IEEE 754 doubles does this, including the spreadsheet on your other screen. Never test two calculated numbers with `==`. Use `near()`, which allows a tiny tolerance:

```{r}
#| label: near
near(0.1 + 0.2, 0.3)
```

Logical values are `TRUE` and `FALSE`, and R treats them as 1 and 0 when you do arithmetic on them. That small fact turns "how many" and "what proportion" into `sum()` and `mean()`:

```{r}
#| label: logical-arith
sum(covid$confirmed > 10000)
mean(covid$stringency_index > 50)
```

254 days of 2020 with more than ten thousand cumulative cases, and the stringency index was above 50 on 57% of days.

### NA

`NA` is R's marker for a value that should exist but is not recorded. It is not zero, not an empty string, and not "the data is fine". It is an admission of ignorance, and R takes it seriously: almost anything you do with an `NA` gives you `NA` back, because R refuses to guess.

```{r}
#| label: na-basics
NA + 1
mean(c(1, 2, NA))
```

That propagation is the point. If you do not know one of three numbers, you do not know their mean, and R would rather tell you so than quietly average the two it has.

The `deaths` column in this file makes it concrete. Singapore recorded no COVID-19 deaths for the first two months of the epidemic, and this file records those days as missing rather than as zero.

```{r}
#| label: na-deaths
head(covid$deaths, 12)
sum(is.na(covid$deaths))
mean(covid$deaths)
```

59 missing days, and so the mean of the column is `NA`. To get an answer you have to say explicitly that you are willing to ignore the gaps:

```{r}
#| label: na-rm
mean(covid$deaths, na.rm = TRUE)
```

Whether that is the right thing to do is a question about deaths, not about R. The chapter on missing data returns to it at length. What matters here is that R made you ask.

Two more things about `NA`. Testing it with `==` does not work, because "an unknown value" is not equal to anything, including another unknown value:

```{r}
#| label: na-equals
NA == NA
```

Use `is.na()` instead:

```{r}
#| label: is-na
is.na(c(1, NA, 3))
```

And `NA` has flavours. Bare `NA` is logical; there are also `NA_integer_`, `NA_real_` and `NA_character_`, so that a missing value can sit inside a vector without changing its type. You rarely type these yourself, but you will see them in error messages.

```{r}
#| label: na-types
typeof(NA)
typeof(NA_character_)
typeof(c(NA, 1))
```

## Coercion, or how a column of numbers becomes text {#sec-coercion}

A vector holds one type. So what happens when you combine two types?

```{r}
#| label: coercion
c(1, "a")
```

R did not complain. It quietly converted the number `1` into the string `"1"`, because that is the only way to keep everything in one vector. The rule runs one way only, from least to most flexible: logical becomes integer becomes double becomes character.

```{r}
#| label: coercion-chain
typeof(c(TRUE, 1L))
typeof(c(1L, 2.5))
typeof(c(2.5, "Singapore"))
```

Nothing here is a bug, and no warning is issued at any step. That is exactly why it costs people an evening. Picture a spreadsheet of daily deaths where one cell says `"not reported"` instead of being left blank. That single cell turns the entire column to character on import, and then:

```{r}
#| label: mean-character
mean(c("1", "a"))
```

The warning is the useful part: `argument is not numeric or logical: returning NA`. When you see it, you have a character column that you thought was numeric. `typeof()` on the column confirms it in one line.

Going the other way is explicit, and R does warn:

```{r}
#| label: as-numeric
as.numeric(c("1", "a"))
```

`NAs introduced by coercion` means R turned something into `NA` because it could not parse it. Sometimes that is what you wanted. Sometimes it has just deleted a fifth of your data. Always check how many `NA`s appeared.

## Recycling {#sec-recycling}

When two vectors in an arithmetic expression have different lengths, R repeats the shorter one until it is long enough. You have already relied on this: in `confirmed * 2`, the `2` is a vector of length one and it got recycled 345 times.

That is the useful case. Here is the other one.

::: {.callout-important}
## Predict before you run

`c(1, 2, 3, 4) + c(10, 20)` produces a vector of length four. Write down the four numbers you expect, and whether you expect a warning. Then run it.
:::

```{r}
#| label: recycling-quiet
c(1, 2, 3, 4) + c(10, 20)
```

No warning at all. Four divides by two, so R considered the recycling clean and said nothing. Change the length to three and it does speak up:

```{r}
#| label: recycling-warn
c(1, 2, 3) + c(10, 20)
```

`longer object length is not a multiple of shorter object length` is R telling you your two vectors do not line up. It is the same silent mechanism either way, and the warning only appears when the arithmetic is untidy, not when it is wrong. This is why `length()` is worth running on anything you are about to combine.

## Calling functions {#sec-functions}

A function has a name, takes arguments in brackets, and gives back a value. `mean(new_cases)` takes one argument and returns one number. Functions nest, and R works from the inside out:

```{r}
#| label: nesting
round(mean(new_cases), 2)
median(new_cases)
```

A mean of 169.85 new cases a day against a median of 40 tells you the distribution is badly skewed, which for this epidemic it is: most of 2020 was quiet, and a few weeks in April and May were not.

Arguments can be given by position or by name. These three calls are identical:

```{r}
#| label: named-args
seq(1, 10, 3)
seq(from = 1, to = 10, by = 3)
seq(by = 3, to = 10, from = 1)
```

Position only works if you remember the order. Names work regardless, and they document the call for whoever reads it next, including you in three weeks. The convention most people settle on is to pass the first argument or two positionally, because `mean(new_cases)` is perfectly clear, and to name everything after that.

Most arguments have defaults, so you only supply the ones you want to change. `args()` shows you what they are:

```{r}
#| label: args
args(round)
args(mean)
```

`round(x, digits = 0)` means that if you do not say otherwise, you get whole numbers. `mean(x, ...)` is less helpful, because `mean` is generic and the real arguments live in the method; this is one of the places where R's help page tells you more than the code does.

While we are here, `round()` has an oddity that catches people out:

```{r}
#| label: banker
round(0.5)
round(1.5)
round(2.5)
```

`round(0.5)` is 0 and `round(1.5)` is 2. R rounds halves to the nearest even number rather than always upward, following the IEC 60559 standard, so that rounding a long column of values does not drift upward. `?round` says so explicitly. It is documented behaviour, not a bug, and it will still confuse a colleague checking your totals by hand.

### Reading a help page

Every function in R ships with documentation. A question mark opens it:

```{r}
#| label: help
#| eval: false
?mean
help("mean")
```

If you do not know the name, two question marks searches the text of every installed help page:

```{r}
#| label: help-search
#| eval: false
??"standard deviation"
```

Help pages have a fixed shape, and reading them in order is a waste of time. Read them in this order instead.

**Usage** first. It shows the function signature with every default filled in, so you can see at a glance what you are allowed to pass. **Arguments** second, but only the two or three you care about. **Examples**, at the very bottom, third and often most useful; they are runnable, and `example(mean)` runs them for you. **Details** only when something has already gone wrong, because that is where the awkward truths live, including the rounding rule above. **Value** when you are unsure what type comes back, which matters more than beginners expect.

Tidyverse functions carry the same documentation, typeset more kindly at <https://tidyverse.org>. If a help page is impenetrable, that is normal rather than a sign you are slow. Much of base R's documentation was written for statisticians in the 1990s and reads that way. Search for the function name plus "r" and you will find people explaining it to each other in ordinary language.

## Packages and where functions live {#sec-packages}

R ships with a few thousand functions. Everything else lives in packages: bundles of functions, documentation and sometimes data, written by other people and published on CRAN.

Installing is a one-off, per machine. It downloads and compiles code onto your hard disk.

```{r}
#| label: install
#| eval: false
install.packages("tidyverse")
```

Loading is per session. It takes an installed package and makes its functions available by name.

```{r}
#| label: library
#| eval: false
library(tidyverse)
```

The distinction trips up everyone once. `install.packages()` belongs at the console, typed by hand, and should never sit in a script; running it on every render wastes minutes and needs the internet. `library()` belongs at the top of every script, because a script that does not load what it uses will not run on anyone else's machine, including yours next month. The quotation marks go on one and not the other, which is an inconsistency in R and not something you did wrong.

### The masking message

The first time you ran `library(tidyverse)` you got a wall of text. Read it rather than scrolling past it.

```
── Attaching core tidyverse packages ──────────────────────── tidyverse 2.0.0 ──
✔ dplyr     1.2.1     ✔ readr     2.2.0
✔ forcats   1.0.1     ✔ stringr   1.6.0
✔ ggplot2   4.0.3     ✔ tibble    3.3.1
✔ lubridate 1.9.5     ✔ tidyr     1.3.2
✔ purrr     1.2.2
── Conflicts ────────────────────────────────────────── tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
ℹ Use the conflicted package (<http://conflicted.r-lib.org/>) to force all conflicts to become errors
```

Load dplyr on its own and it says the same thing more bluntly:

```
Attaching package: ‘dplyr’

The following objects are masked from ‘package:stats’:

    filter, lag

The following objects are masked from ‘package:base’:

    intersect, setdiff, setequal, union
```

Nothing has broken. Here is the mechanism. R keeps an ordered list of places to look for a name, called the search path:

```{r}
#| label: search-path
head(search(), 15)
```

When you type `filter(...)`, R walks that list from left to right and uses the first `filter` it finds. Every package you attach is inserted near the front, so a newly loaded package wins against anything loaded before it. `package:dplyr` sits ahead of `package:stats`, so `filter` means dplyr's.

`stats::filter` has not been deleted, and you can still reach it by giving its full address. The double colon means "the function called `filter` from the package called `stats`, regardless of the search path":

```{r}
#| label: stats-filter
head(round(stats::filter(covid$confirmed, rep(1/7, 7)), 1), 10)
```

That is a seven-day moving average of the cumulative curve, which is what `stats::filter()` is for. It returns a time series object, hence the header lines about start, end and frequency, and the first three values are `NA` because a seven-day window cannot be centred on day one. None of it has anything to do with picking rows out of a table:

```{r}
#| label: dplyr-filter
dplyr::filter(covid, stringency_index > 80)$date
```

Two functions, one name, completely different jobs. The message exists so that you find out about the collision when you load the package rather than three hours later when a result looks wrong.

The same applies to `lag()`. `dplyr::lag()` shifts a vector along by one position, which is exactly what you did by hand earlier. `stats::lag()` shifts a time series' *index* and leaves the values alone. If you ever compute daily cases and get a column of zeroes, this is why.

The practical consequence, and the reason `library()` goes at the top of a script: if you forget to load dplyr and call `filter()` on a data frame, you get `stats::filter`, and the error tells you nothing about the real problem.

```
Error: object 'confirmed' not found
```

`stats::filter()` has no idea that `confirmed` is a column of the thing you handed it, so it tries to find an object by that name and fails. The fix is a `library(dplyr)` or `library(tidyverse)` line you left out, and the error message does not say so.

You can sidestep the whole business by writing `dplyr::filter()` in full. That is worth doing in a function you are shipping to other people, and it is too noisy for everyday analysis. Load the package, read the conflicts message, and if you want the collisions to become hard errors instead of quiet precedence, the `conflicted` package does that.

## Exercise

::: {#exr-vectors}
## Build a week by hand

The first seven cumulative confirmed counts in the file are 0, 1, 3, 3, 4, 5 and 7.

1. Store them in a vector called `week_one` and confirm that R agrees it has seven elements and holds doubles.
2. Work out how many new cases were recorded on each of days two to seven, using only subsetting and subtraction.
3. Add the label `"week one"` to the front of `week_one`, look at the type again, and then try to multiply the result by two. Explain in one sentence what R did to your numbers and why the arithmetic now fails.
4. Make a copy of `week_one` with the fourth element replaced by `NA`, then find its mean twice, once as R gives it to you and once telling `mean()` to drop the missing value.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

For part 2, two vectors of the same length subtract element by element, so you need one vector holding days two to seven and another holding days one to six. For part 3, look at `typeof()` before and after.
:::

[Worked solution](solutions.qmd#sec-sol-vectors)
