# Good practice in a nutshell

```{r}
#| label: setup
#| include: false
source("_common.R")
```

A large share of the time beginners lose in R goes on workflow: the file that was in the wrong folder, the object left over from an hour ago that changed a result without any visible sign, and the package that changed under a script nobody had run in a year. None of these are difficult problems, but they are much easier to avoid from the start than to correct in an established project.

This chapter sets out ten working habits that avoid them.

## 1. Work in a project

Make a folder for each piece of work, put a `.Rproj` file in it, and open your work by double-clicking that file.

A `.Rproj` is a small plain-text file. Here is the whole of the one at the root of this book, `r-intro.Rproj`, with nothing removed:

```
Version: 1.0

RestoreWorkspace: No
SaveWorkspace: No
AlwaysSaveHistory: No

EnableCodeIndexing: Yes
UseSpacesForTab: Yes
NumSpacesForTab: 2
Encoding: UTF-8

RnwWeave: knitr
LaTeX: XeLaTeX

AutoAppendNewline: Yes
StripTrailingWhitespace: Yes
```

That is the complete file. Its effects are worth listing. Opening it starts a fresh R session whose working directory is the folder containing the file, on any machine, without a path being typed. It gives the project its own environment and its own command history, so two projects open at the same time do not share objects. It carries the settings above with the folder, so a collaborator who opens it gets the project's tab width and encoding rather than their own. And it marks the folder: several packages, including `here`, `renv` and `usethis`, locate the project root by looking for a `.Rproj` file.

Make one with File, New Project, Existing Directory. It takes a few seconds and the remaining rules in this chapter assume it.

## 2. Use `here()` rather than `setwd()`

A script can be pointed at its data with `setwd()` and an absolute path:

```{r}
#| label: setwd-trap
#| eval: false
setwd("/Users/swapnil/Documents/GitHub/r-intro")

covid <- read_csv("data/sg_covid_data.csv")
```

The path is specific to one machine and to the current arrangement of the folders on it, so the script fails for a colleague and for you after any reorganisation. On another machine the first line produces:

```
Error in setwd("/Users/swapnil/Documents/GitHub/r-intro") :
  cannot change working directory
```

No single absolute path is correct on two machines with different users or directory layouts. Windows adds a second problem, because Windows paths use backslashes and a backslash begins an escape sequence in an R string:

```
Error: '\U' used without hex digits in character string
```

That is R reading `C:\Users\...`, seeing `\U`, and expecting a Unicode code point to follow. You can escape the backslashes, or use forward slashes, which R accepts on Windows. The better option is to avoid absolute paths.

`here::here()` builds paths without an absolute component being typed. It walks up from the current working directory looking for a project marker, which may be a `.Rproj` file, a `.git` directory, or a file called `.here`, and then builds a path down from the folder it finds:

```{r}
#| label: here-demo
here::here("data", "sg_covid_data.csv")
```

That is the path on the machine that built this book; on yours the same call prints a different absolute path but points at the same file within the project. The call also behaves the same way from a script in the project root, from a Quarto document two folders down, and from the console, because it resolves from the project root rather than from the current working directory.

You give `here()` the path components as separate arguments and it joins them with the separator the operating system uses. `here("data", "raw", "file.csv")` is therefore correct on macOS, Linux and Windows without a slash being typed in either direction.

## 3. Restart R often, and start each session empty

This rule involves two settings, changed once, and one keyboard shortcut.

Go to Tools, Global Options, General. Untick **"Restore .RData into workspace at startup"**. Set **"Save workspace to .RData on exit"** to **Never**. By default, restore is ticked and save is set to "Ask", so the workspace is preserved unless you decline the dialog each time; setting it to Never removes the question. Both defaults date from a period when reloading a saved workspace was cheaper than recomputing it, and RStudio retains them for backward compatibility.

The `.Rproj` file overrides the global setting for its own project, which is why the one shown above carries `RestoreWorkspace: No` and `SaveWorkspace: No`. Change both in Global Options as well, so that projects you have not configured behave the same way.

Then learn <kbd>Ctrl</kbd>+<kbd>Shift</kbd>+<kbd>F10</kbd>, or <kbd>Cmd</kbd>+<kbd>Shift</kbd>+<kbd>F10</kbd> on a Mac, which restarts R. Use it several times an hour. It takes under a second.

The failure this prevents is as follows. When R saves the workspace on exit and reloads it on startup, the environment accumulates. Objects created weeks ago remain, and are not visible unless you look in the Environment pane. A script can then run without error on your machine and fail on line 12 for someone else, because line 12 uses an object the script never creates and which has been sitting in your workspace since March. The script is incomplete, and the accumulated environment conceals the omission.

A second version of the problem is harder to detect, and this one comes from a bug in the worksheet this book replaces. The worksheet had a loop that built a monthly subset in order to print a count for each month, close to this:

```{r}
#| label: stale-loop
covid <- read_covid() %>%
  arrange(date) %>%
  mutate(daily_cases = confirmed - lag(confirmed, default = 0))

for (m in 1:12) {
  month_data <- covid %>% filter(month(date) == m)
  # the original printed a count for month m here
}
```

The loop runs and finishes. Because `m` and `month_data` were created at top level, they remain in the global environment afterwards, holding whatever the last iteration put in them:

```{r}
#| label: stale-leftovers
m
range(month_data$date)
```

The subset left behind is December's. Forty lines further down the worksheet, after several other exercises, there was a plot of `month_data` that was meant to show April, because the reader had been asked to build an April subset in the meantime. If they skipped that exercise, or ran it and then restarted, or ran the chunks out of order, `month_data` was still December and the plot was plausible on its face.

```{r}
#| label: stale-consequence
# What the reader thought was April's mean daily cases
mean(month_data$daily_cases)
```

```{r}
#| label: stale-truth
# What April actually was
covid %>%
  filter(month(date) == 4) %>%
  summarise(april_mean = mean(daily_cases))
```

The stale object gives a mean of twelve daily cases; April's mean is five hundred and eight. Both print without complaint and nothing raises an error. A stale global object is difficult to detect for exactly that reason: there is no warning, and the only symptom is a number that happens to be wrong.

Restarting R and running the script from the top detects this. If the script runs from an empty session, its results depend only on what it contains. If it does not, you have found a dependency on something outside the file.

## 4. Loading packages, and clearing the environment

Put every package your script needs in one block at the top of the file. A reader can then see the dependencies without reading the whole script. A script that loads a package halfway down also fails differently depending on how far execution reached before the failure, which makes the failure harder to reproduce.

There is a one-line alternative to restarting, which empties the global environment:

```{r}
#| label: rm-trap
#| eval: false
rm(list = ls())
```

This is not equivalent to a restart. `ls()` lists the objects in the global environment and `rm()` removes them; nothing else changes. Several things are left untouched.

Attached packages stay attached. `search()` shows R's chain of environments, and `rm(list = ls())` empties one entry in it, the first:

```{r}
#| label: search-path
search()
```

Everything from `package:scales` down is still attached, and still masks whatever it was masking. Session-wide settings survive as well: an `options(digits = 3)` set an hour ago continues to round printed output. Open file connections stay open. And `ls()` omits names beginning with a dot, so dot-named objects are never passed to `rm()`:

```{r}
#| label: dot-names
set.seed(42)
x <- runif(1)
rm(list = ls())

ls()
ls(all.names = TRUE)
```

`.Random.seed` survives the clean-out, so the random number state carries over as well. Restarting R clears all of this, and takes about as long as typing the `rm()` call.

## 5. Treat the raw data file as read-only

Whatever arrives from the source goes in `data/` and is not edited afterwards: not to fix a date format, not to fill in a blank, not to remove a duplicated header row.

The dataset in this book illustrates the reason. It has 59 missing values at the start of the deaths column, 24 missing days in each of the six mobility columns, and case counts that are cumulative rather than daily. All three need handling, and each is handled in code that you can read and disagree with. Opening the CSV in Excel and typing zeros over the missing deaths would produce a tidier-looking file and a worse analysis, because a reported zero and a not-yet-reported value are different quantities and the edited file would no longer distinguish them.

Repairs made in the script carry their justification with them, survive the arrival of an updated version of the source file, and can be inspected by a reviewer. An edit made by hand in a spreadsheet has none of those properties.

## 6. Naming objects

Use `snake_case`: lowercase words joined by underscores. `daily_cases`, `monthly_summary`, `covid_april`. At least four naming conventions coexist in R's base installation, so the language itself will not be consistent, but your own code can be.

Name data after what it is, using nouns, and functions after what they do, using verbs. `severity` is a column; `classify_severity()` is a function. In `monthly_report <- summarise_by_month(covid)` the distinction is visible without consulting anything.

Avoid `df`, `df2`, `data`, `temp` and `df_final_v2`, for two reasons. The first is that `df2` carries no information about its contents, so six weeks later you have to run the code to find out what is in it. The second is that some of those names are already in use. `df` is a function in base R, the density of the F distribution:

```{r}
#| label: df-is-a-function
df(1, 2, 3)
```

Assigning a data frame to `df` masks that function. This particular collision rarely causes trouble, since few analyses call `df()`. Others are riskier: `data`, `c`, `mean`, `sum`, `t` and `T` are all short, tempting names that already belong to something in base R, and code that calls the masked function afterwards will fail or, in the case of `T`, silently change meaning.

Long names are cheap to type, because Tab completes them. `monthly_case_summary` is a few keystrokes and a Tab.

## 7. Pin your packages with `renv`

`renv` gives a project its own package library and a lockfile recording the exact version of every package in it. Three commands cover ordinary use:

```{r}
#| label: renv-commands
#| eval: false
renv::init()      # once, when you start the project
renv::snapshot()  # after you add or update a package
renv::restore()   # on a new machine, or when you come back to it
```

The claim a lockfile supports is a narrow one: the repository together with `renv.lock` records which package versions produced the results, and `renv::restore()` reinstalls those versions. Without a lockfile, "install the tidyverse and run my script" means whatever the tidyverse happens to be on the day someone tries, and package behaviour does change between versions.

This book uses `renv`, and `renv.lock` is committed alongside the chapters. If you clone the repository, `renv::restore()` builds the library the book was written against.

The limits are worth stating. `renv` pins R packages. It does not pin R itself, the operating system, or the system libraries a package compiles against, and reproducing those requires containers, which are outside the scope of this book. Pinning packages addresses the largest and most frequent source of version drift.

## 8. Version control with Git

```bash
git add -A
git commit -m "Compute daily cases from the cumulative column"
git push
```

`add` chooses what goes into the next snapshot, `commit` takes the snapshot with a message attached, and `push` sends it to a machine that is not your laptop. Those three are enough to be useful from the start; the rest of Git can wait until a situation calls for it.

The property this gives you is a record of the project that is difficult to alter unintentionally. Every commit is a complete state of the project with a timestamp, an author and a message, and once pushed it also exists on a server. History can be rewritten deliberately, but not by accident, and rewriting a commit someone else has already pulled is visible to them. Saving successive copies of a script under new filenames records none of that: no timestamped author, no message, and no way to tell which copy produced which result.

The everyday benefit is more immediate. With commits behind you, code can be deleted rather than kept: the half-finished approach and the two hundred lines you replaced can both go, the file stays readable, and any of it can be recovered with one command. Without version control the safer option is to keep everything, and scripts grow accordingly.

Commit code, small data files, and your lockfile. Do not commit two gigabytes of intermediate output. That is what `.gitignore` is for.

## 9. Ask for help with a reprex

A **reprex** is a reproducible example: the smallest piece of code that still shows your problem, together with everything someone needs to run it. The `reprex` package will build one for you. Copy your code, run `reprex::reprex()`, and it executes that code in a clean R session and puts the code plus its real output on your clipboard, formatted for GitHub or Slack.

The clean session does useful work here. A fair proportion of the time, writing the reprex resolves the problem without anyone else reading it, because the cause was an object left over in your environment which the clean session does not have. This is the same failure described in rule 3.

What makes a question answerable, in order of how often it is missing:

1. Code the reader can run. A screenshot cannot be run, and neither can a fragment referring to `my_data`, which exists only in your session. Either use a dataset that ships with R, or reduce yours to ten rows and paste the output of `dput(head(your_data, 10))`.
2. The error message, copied exactly rather than paraphrased or retyped from memory, including the lines after the first.
3. What you expected instead. Stating the expectation exposes any disagreement about what the function was supposed to do, which is often where the problem actually lies.
4. Your versions. `sessionInfo()` prints your R version and every attached package. Paste it at the bottom.

Remove anything not needed to reproduce the problem. If the example is forty lines and the error occurs on line 4, the other thirty-six lines have to be read and discarded by whoever answers.

## 10. Write the report and the analysis in the same file

Running the script, copying the numbers into Word and writing around them leaves the document and the analysis as two separate artefacts, and they agree only until the analysis changes. Change a filter, re-run the script, and the number in the document is out of date with nothing to indicate it.

Quarto removes the copying step. A `.qmd` file holds prose and code together, and rendering it runs the code and inserts the resulting output into the finished document. Change the filter, render again, and the numbers update: tables, figures, and values written inline in a sentence alike.

This book is a Quarto book. Every chapter is a `.qmd` file, every output on every page was produced by the code shown above it, and the book is rebuilt from scratch before publication, so a number in the text cannot drift away from the code that produced it.

## The checklist

The ten rules above, in one screen.

| | |
|---|---|
| **Project** | One folder per piece of work, with a `.Rproj` in it. Open by double-clicking that file. |
| **Paths** | No `setwd()`. No absolute paths. `here::here("data", "file.csv")`. |
| **Session** | Save workspace: **Never**. Restore `.RData`: **never**. Restart R (<kbd>Ctrl/Cmd</kbd>+<kbd>Shift</kbd>+<kbd>F10</kbd>) several times an hour. |
| **Test** | Before you send it: restart R, run the whole script from the top. If it does not run cleanly, something it needs is missing from the file. |
| **Packages** | All `library()` calls in one block at the top. `rm(list = ls())` is not a restart. |
| **Data** | `data/` is read-only. Every repair happens in code. |
| **Names** | `snake_case`. Nouns for data, verbs for functions. Avoid `df`, `df2`, `temp`. |
| **Versions** | `renv::init()` at the start, `renv::snapshot()` when you add a package, commit `renv.lock`. |
| **History** | `git add -A`, `git commit -m "..."`, `git push`. Commit early, delete freely. |
| **Help** | `reprex::reprex()`. Runnable code, the exact error, your expectation, `sessionInfo()`. |
| **Output** | Report and analysis in one `.qmd`. No numbers pasted into a separate document. |

: The ten habits in this chapter. {tbl-colwidths="[18,82]"}
