# Grouping and summarising {#sec-groups}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

```{r}
#| label: data
#| include: false
covid <- covid_daily()
```

Almost nothing you are asked for in public health is a number about the whole dataset. It is a number per month, per ward, per age band, per vaccination status. "How many cases were there in 2020" is a trivia question. "How did the monthly average move across the circuit breaker" is an actual question, and answering it means splitting 345 rows into piles, computing something in each pile, and gluing the answers back into a table.

dplyr calls that split-apply-combine, and it is two verbs: `group_by()` to say what the piles are, `summarise()` to say what to compute in each. Before either of them, though, there is `count()`, which does the most common case in one word.

Every chunk below uses `covid`, the dataset you finished the last chapter with. In your own session, `covid <- covid_daily()` rebuilds it: daily counts differenced out of the cumulative columns, `severity` as an ordered factor, and a `month` column.

## `count()` does the obvious thing

```{r}
covid %>% count(severity)
```

215 days Low, 53 Medium, 77 High. The column of counts is called `n`, which is dplyr's convention everywhere.

`sort = TRUE` puts the biggest pile first, which is what you want when there are forty categories and you care about the top few:

```{r}
covid %>% count(severity, sort = TRUE)
```

Give it two columns and it counts every combination that occurs:

```{r}
covid %>% count(month, severity)
```

Combinations that never happened are absent rather than zero. January has one row, not three, because there were no Medium or High days in January. That is `count()` behaving reasonably and it is also a trap: if you feed this straight into a plot, the missing zeroes leave holes. `count(month, severity, .drop = FALSE)` keeps the empty factor levels when you need a complete grid.

## `group_by()` and `summarise()`

`count()` is a shorthand. The long form spells out both halves, and once you can write the long form you can compute anything, not only counts.

```{r}
covid %>%
  group_by(severity) %>%
  summarise(
    days       = n(),
    mean_cases = mean(daily_cases),
    worst      = max(daily_cases)
  )
```

`group_by()` changes nothing about the data. It attaches a note saying "treat this as three piles". `summarise()` reads the note, runs each expression once per pile, and returns one row per pile. The 345-row dataset becomes a 3-row table, and every column in the output is something you asked for by name.

`n()` is special: it takes no arguments and returns the number of rows in the current group. It only works inside dplyr verbs, because outside one there is no "current group" for it to count.

### Counting with `sum()`

Here is the idiom worth memorising. In R, `TRUE` is 1 and `FALSE` is 0 when you do arithmetic on them. So `sum()` of a condition counts how many times it was true:

```{r}
covid %>%
  summarise(
    n_high      = sum(severity == "High"),
    n_zero_days = sum(daily_cases == 0),
    prop_high   = mean(severity == "High")
  )
```

77 High days, 13 days with no new cases at all, and High days were 22% of the year. `mean()` of the same condition gives the proportion, for the same reason: the average of a pile of ones and zeros is the fraction that are ones.

This idiom composes with grouping, which is where it earns its keep, and it saves you from the awkward alternative of filtering and then counting rows. One warning, and it will bite you on this dataset:

```{r}
covid %>%
  summarise(
    death_days = sum(daily_deaths > 0)
  )
```

`NA`. The deaths column is missing for the first 59 days of 2020, so 60 of the comparisons return `NA`, and a sum containing `NA` is `NA`. R is telling you it cannot answer, which is correct: it does not know whether those days had deaths. Add `na.rm = TRUE` only once you have decided that treating "not reported" as "no deaths" is defensible for what you are writing.

```{r}
covid %>%
  summarise(
    death_days = sum(daily_deaths > 0, na.rm = TRUE)
  )
```

25 days on which the cumulative death total went up. Not 29 deaths, and not 27 either. Those are three different quantities and the distinction is the previous chapter's problem, not this one's.

### The `.groups` message

Group by two things and `summarise()` says something on the way out:

```{r}
monthly_severity <- covid %>%
  group_by(month, severity) %>%
  summarise(days = n())

monthly_severity
```

Read it rather than fleeing from it. You grouped by two variables, `summarise()` consumed the last one, and the result is still grouped by `month`. That matters, because the next verb in your pipeline will operate per month whether or not you wanted it to. Nothing has gone wrong; dplyr is warning you that the object in your hand is not a plain table.

State your intention and the message goes away:

```{r}
covid %>%
  group_by(month, severity) %>%
  summarise(days = n(), .groups = "drop")
```

`.groups = "drop"` returns an ungrouped table. `.groups = "drop_last"` keeps the old behaviour explicitly. Use `"drop"` unless you have a reason not to, and reach for `ungroup()` if you inherit a grouped object from somewhere else. Leftover grouping is one of the most common sources of a number that is right in your head and wrong on the screen.

## Grouped `mutate()` keeps every row

`summarise()` collapses. `mutate()` does not. Group first and `mutate()` still returns one row per input row, but the calculations inside it see only the current group. That is how you compare each day with its own month.

::: {.callout-important}
## Predict before you run

August 2020 had 31 days and 4,607 cases. The next chunk works out what share of the month's total fell on its single busiest day. Write down the percentage you expect. An even spread would put each day at about 3%.
:::

```{r}
covid %>%
  group_by(month) %>%
  mutate(share_of_month = daily_cases / sum(daily_cases)) %>%
  ungroup() %>%
  filter(month == as.Date("2020-08-01")) %>%
  slice_max(share_of_month, n = 3) %>%
  select(date, daily_cases, share_of_month)
```

`sum(daily_cases)` inside a grouped `mutate()` is August's total, repeated down every August row, so the division gives each day's share of its own month. If you had run the same `sum()` without grouping you would have divided by 58,599 and got a very small number for everything.

Look at what it found. 5 August 2020 alone accounts for nearly a fifth of the month. August's median day was 91 cases and the next highest day was 313, so 908 is not the shape of an epidemic curve. MOH's account at the time attributed the bulk of it to clearance testing in the dormitories: 903 of the 908 were dormitory residents and 581 were linked to a single cluster at Acacia Lodge, many of them past infections detected by that testing rather than new transmission on the day. A monthly mean will fold that spike straight into August's average without comment, which is why a grouped `mutate()` that shows you each day against its own month is a diagnostic worth running before you summarise anything.

The rule of thumb: use `summarise()` when the answer has fewer rows than the input, and a grouped `mutate()` when the answer has the same number of rows but each one needs to know something about its neighbours.

```{r}
nrow(covid %>% group_by(month) %>% mutate(total = sum(daily_cases)))
nrow(covid %>% group_by(month) %>% summarise(total = sum(daily_cases)))
```

## Grouping by month, done twice

The `month` column exists because `covid_daily()` ran `floor_date(date, "month")`. Rounding a date down to the start of its month is a lubridate one-liner, and the result is still a `Date`:

```{r}
head(floor_date(covid$date, "month"), 3)
class(floor_date(covid$date, "month"))
```

`floor_date()` takes any unit you like, so the same idea gives you weeks or quarters without touching the code that follows:

```{r}
head(floor_date(covid$date, "week"), 3)
head(floor_date(covid$date, "quarter"), 3)
```

The worksheet this book replaces did it a different way, by cutting the first seven characters out of the date:

```{r}
covid %>%
  mutate(month_chr = substr(date, 1, 7)) %>%
  count(month_chr) %>%
  head(3)
```

That works. It gives the right twelve groups in the right order and, for a one-off count, nobody would be harmed. But look at what it produced:

```{r}
covid %>%
  mutate(month_chr = substr(date, 1, 7)) %>%
  pull(month_chr) %>%
  class()
```

`substr()` has no idea what a `Date` is. It calls `as.character()` on the way in, chops the text, and hands you back text. Your grouping variable is now a string that happens to look like a date, and three things follow.

The ordering is correct by luck. ISO-8601 strings sort chronologically when sorted alphabetically, which is the entire reason the format was designed that way, and it is why nobody notices this problem on their first dataset. Read the file in with a different date format and the luck runs out silently:

```{r}
uk_style <- format(covid$date, "%d/%m/%Y")
head(uk_style, 3)
head(substr(uk_style, 1, 7), 3)
sort(unique(substr(uk_style, 1, 7)))[1:5]
```

Seven characters of `22/01/2020` is `22/01/2`, and sorting those puts every 1st of the month first regardless of year. No error. A plot with a wrong x-axis.

Arithmetic is gone. A `Date` knows how to add a month or measure a gap; a string does not:

```{r}
#| error: true
head(substr(covid$date, 1, 7)) + months(1)
```

```{r}
head(covid$month + months(1), 3)
max(covid$month) - min(covid$month)
```

And ggplot will treat the character column as an unordered category, so you lose date axes, sensible breaks and the ability to put a vertical line at 7 April 2020 where the circuit breaker started.

You cannot rescue it by parsing the string back, either, because a year-month is not a date:

```{r}
#| error: true
as.Date("2020-09")
```

**character string is not in a standard unambiguous format**. It errors rather than returning `NA`, which means a pipeline built on partial date strings falls over rather than producing quiet rubbish. That is the better failure of the two, but you should not be relying on either. Round the date, keep it a date.

## Months are not the same length

Here is the monthly table, with the thing everybody forgets included from the start:

```{r}
monthly <- covid %>%
  group_by(month) %>%
  summarise(
    days       = n(),
    cases      = sum(daily_cases),
    mean_daily = mean(daily_cases),
    high_days  = sum(severity == "High"),
    .groups    = "drop"
  )

monthly
```

January has 10 days. The data starts on 22 January, so "January" here is not a month, it is the last third of one. February has 29 because 2020 was a leap year. Comparing 13 cases in January with 89 in February and concluding the epidemic grew sevenfold would be wrong twice over: wrong on the denominator, and wrong on what the numbers even are.

The `days` column is the fix, and it costs one `n()` call. Whenever you total something over a period, put the period length in the table next to it. A per-day rate is comparable; a monthly total is not.

```{r}
#| label: fig-monthly-mean
#| fig-cap: "Mean new cases per day, by month. January covers 10 days and is drawn on the same footing as the rest only because the average divides by its own denominator."
ggplot(monthly, aes(x = month, y = mean_daily)) +
  geom_col() +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(x = NULL, y = "Mean new cases per day")
```

## The report is a data frame

Now the exercise the original worksheet set: flag every month with more than five High-severity days. The rule itself is one line of the last chapter's material.

```{r}
alerts <- monthly %>%
  mutate(status = if_else(high_days > 5, "Alert", "All clear"))
```

The worksheet's answer then looped over the months and printed a line each time. Its output looked like this:

```{r}
for (i in 1:3) {
  cat(format(alerts$month[i], "%B"), "-", alerts$status[i], "\n")
}
```

The loop itself is doing something simple: `i` takes the values 1, 2 and 3 in turn, and `cat()` pastes its arguments together and prints them to the console. Look at what you are left with holding. Those three lines are text on a console. You cannot sort them, filter them, join them to a table of policy dates, count how many said "Alert", or paste them into a document without retyping. When the loop finishes, the answer is gone.

The table is the report:

```{r}
alerts
```

Same information, and now it is an object. Sorting it, filtering it, writing it to CSV and plotting it are all one more line each:

```{r}
alerts %>% filter(status == "Alert")
```

Four months, April through July, which is the dormitory outbreak and the circuit-breaker period around it. August drops below the threshold with 4 High days despite containing the 908-case spike, because the rule counts days rather than cases and one enormous day is still one day. Whether that is the right rule depends on what the alert is for, and noticing that it is a choice is more than half of what this exercise teaches.

You still need sentences sometimes, for an email or a figure caption. `str_glue()` builds one by substituting anything in curly braces:

```{r}
worst <- monthly %>% slice_max(cases, n = 1)

str_glue("The worst month was {format(worst$month, '%B %Y')}: {worst$cases} cases over {worst$days} days.")
```

Build the table first and glue a sentence out of it at the end. Going the other way, where the sentence is the only artefact, is how you end up recomputing the whole analysis because someone asked a follow-up question.

## Checking September against something else

The original worksheet asked for September's mean daily cases twice: once with `mean()`, and once with a loop that added the values up and divided by 30. Both gave the same answer, and both would have gone on giving the same answer if `daily_cases` had been computed wrongly, because they were the same arithmetic on the same column. A check that shares its inputs with the thing it is checking is not a check.

Here is one that is. `confirmed` is cumulative and untouched by us. September's new cases must therefore equal the cumulative total on 30 September minus the total on 31 August, whatever our `lag()` did or did not do. If `daily_cases` were broken, the two would disagree.

```{r}
september <- covid %>% filter(month == as.Date("2020-09-01"))

from_daily <- september %>%
  summarise(days = n(), cases = sum(daily_cases), mean_daily = mean(daily_cases))

from_daily
```

```{r}
bounds <- covid %>%
  filter(date %in% as.Date(c("2020-08-31", "2020-09-30"))) %>%
  select(date, confirmed)

bounds

(bounds$confirmed[2] - bounds$confirmed[1]) / 30
```

31.77 both ways, from two routes that share no intermediate column. The differenced route sums 30 numbers we manufactured. The telescoping route touches the raw cumulative column at exactly two rows and ignores everything in between. Agreement between them is evidence; agreement between `mean()` and a loop doing the same division is not.

The general move: find a quantity you can compute from the untouched source and compare it against the same quantity computed from your derived columns. It takes two lines and it catches the class of error that is otherwise invisible, the one where every number is plausible and all of them are wrong.

September's 30 days averaged 31.8 new cases with a maximum of 86, in a year whose peak day was 1,426. By the end of the third quarter the dormitory outbreak was over.

## Exercises

::: {#exr-monthly-high}
## High-severity days by month

Produce a table with one row per month showing the number of High-severity days, the number of Medium days, and the number of Low days, plus the number of days of data the month contains.

Then answer two questions in a comment. Which month had the most High days, and is that the same month that had the most cases? If they differ, explain in one sentence why they can.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

You can get there with `count(month, severity)` followed by `pivot_wider()`, or with a single `summarise()` containing three `sum()` calls of the form `sum(severity == "High")`. The second route is fewer new ideas and puts the columns in the order you choose.
:::

[Worked solution](solutions.qmd#sec-sol-monthly-high)

::: {#exr-monthly-alert}
## A monthly alert table you can defend

Build a monthly table that assigns each month one of three statuses rather than two: "Alert" when the month has more than 5 High days, "Watch" when it has 1 to 5, and "All clear" when it has none. Include the number of days in each month and the mean daily cases.

Then change the rule so that it triggers on the mean daily case rate rather than on a count of days, using a threshold you pick and state. Report which months change status, and write one sentence saying which of the two rules you would rather defend in a meeting, and why.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

Three statuses means `case_when()` rather than `if_else()`, and the same first-match-wins ordering rules from the previous chapter apply. Keep both status columns in the same table so the comparison is a single `filter()` on the rows where they disagree.
:::

[Worked solution](solutions.qmd#sec-sol-monthly-alert)

::: {#exr-september}
## Cross-check a month that is not September

Pick a month other than September. Compute its mean daily cases from `daily_cases`, then compute the same figure by telescoping the `confirmed` column between the last day of the previous month and the last day of your chosen month. Confirm they agree.

Then try the same trick on January, and explain what goes wrong and why. Fix it, and say what the fix assumes.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

January's problem is that there is no "last day of the previous month" in the file. Think about what `confirmed` was on 22 January 2020 and whether that value lets you avoid needing one.
:::

[Worked solution](solutions.qmd#sec-sol-september)
