# Reading the epidemic in {#sec-import}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

Everything in this book runs on one file: `sg_covid_data.csv`, 345 rows covering 22 January to 31 December 2020, one row per day. Up to now `read_covid()` has been handing it to you. This chapter opens the lid.

Reading a file in is where more analyses go wrong than anywhere else, and the failures are quiet ones. A date column arrives as text and your time series plots in alphabetical order. A missing value coded as `-99` arrives as a number and your mean is nonsense. A leading zero disappears from a postcode. None of these announce themselves. The defence is to read what R tells you at import time instead of scrolling past it, which is why this book leaves import messages switched on.

## Where the file is {#sec-paths}

R has a working directory: the folder it treats as "here" when you hand it a relative path. The chapter on good practice told you never to write `setwd()` in a script, and this is the payoff for obeying that. A script containing `setwd("/Users/swapnil/Documents/GitHub/r-intro")` runs on exactly one computer. Give it to a colleague, or move the folder, or open it on the machine in the teaching lab, and it fails on line one.

`here::here()` solves it. The `here` package looks upward from wherever R happens to be until it finds the top of your project, identified by an `.Rproj` file, a `.git` directory or a `_quarto.yml`, and builds paths from there.

```{r}
#| label: here-root
here()
```

That is this project's root on the machine that rendered this page; on yours it will be a different string, which is the entire point. Ask it for a file by naming the folders as separate arguments and it assembles a path that works on Windows, macOS and Linux without you thinking about which way the slashes lean.

```{r}
#| label: here-path
here("data", "sg_covid_data.csv")
```

Write `here("data", "sg_covid_data.csv")` in every script from now on. It costs a handful of extra characters and removes a whole category of Monday-morning email.

## `read_csv()` {#sec-read-csv}

readr's `read_csv()` takes a path and returns a tibble.

```{r}
#| label: read
covid <- read_csv(here("data", "sg_covid_data.csv"))
```

That block of text is a message rather than an error, and it repays reading line by line.

`Rows: 345 Columns: 13`. Check this against what you expect before you do anything else. If you were told the file covers a year of daily data and R reports 345 rows, that is 2020 from 22 January onward, and it adds up. If R had said 344, you would want to know which day went missing and why.

`Delimiter: ","`. readr worked out the separator by looking at the first few lines. Files exported from European locales often use a semicolon, because the comma is doing duty as a decimal point; `read_csv2()` handles those.

`chr (1)`, `dbl (11)`, `date (1)`. These are the guessed column types, and this is the part people skip. Three of them matter here.

**`date` came in as a real `Date`, not as text.** readr recognises the ISO 8601 form `YYYY-MM-DD` and parses it. That single fact is why the rest of this book can subtract one date from another, sort chronologically, and hand dates straight to `ggplot()` without ceremony.

**Every count is `dbl`, not integer.** readr's guesser does not distinguish whole numbers from decimals; anything numeric arrives as a double. Nothing goes wrong as a result, and it is worth knowing when you wonder why `confirmed` is `<dbl>` when it obviously counts people.

**One column is `chr`.** That is `administrative_area_level_1`, the constant "Singapore" you met in @sec-tidy.

The two lines at the bottom are hints rather than complaints. `spec()` prints the full specification, one line per column, and it is the thing to look at when the summary is truncated with a `...`:

```{r}
#| label: spec
spec(covid)
```

The empty `format = ""` on `date` means readr used its own ISO guesser rather than a format you supplied. The other hint offers you `show_col_types = FALSE` to silence the message. Take that offer only once you have read the message and pinned the types yourself. Silencing a check you have not performed is not the same as performing it.

### The dates readr cannot guess

readr guesses dates only for the ISO form. Hand it anything else and you get character, without a warning, because a column of text is a perfectly valid column of text.

```{r}
#| label: date-guess
read_csv(I("date,cases\n07/04/2020,106\n08/04/2020,142\n"))
```

`<chr>`. And the trap is that `07/04/2020` is 7 April in Singapore and 4 July in Chicago, so nothing but your knowledge of the source can settle it. Tell readr the format explicitly with the same `%d`, `%m`, `%Y` codes that `strptime()` uses:

```{r}
#| label: date-format
read_csv(
  I("date,cases\n07/04/2020,106\n08/04/2020,142\n"),
  col_types = cols(
    date  = col_date(format = "%d/%m/%Y"),
    cases = col_double()
  )
)
```

### Missing values that are not blank

`read_csv()` treats an empty cell and the literal string `NA` as missing. Health data rarely obliges. If your file codes missing as `-99`, `.`, `N/A` or `Not reported`, say so, or those values will be read as data:

```{r}
#| label: na-arg
#| eval: false
read_csv(
  here("data", "some_other_file.csv"),
  na = c("", "NA", "N/A", "Not reported")
)
```

Get this wrong with a numeric sentinel such as `-99` and nothing will look broken. Your minimum will be `-99`, your mean will be dragged down, and every graph will still render. Check the minimum and maximum of every numeric column after import, every time.

## What `read.csv()` would have given you {#sec-base-read-csv}

Base R has its own reader, spelled with a dot instead of an underscore. Most older course notes use it, so you need to recognise it. The two names are close enough that you will type the wrong one at some point.

```{r}
#| label: base-read
old <- read.csv(here("data", "sg_covid_data.csv"))
class(old)
class(old$date)
head(old$date, 3)
```

No message at all, a plain `data.frame` rather than a tibble, and `date` is a character vector of strings that happen to look like dates. R will not stop you doing something silly with them:

```{r}
#| label: date-arithmetic
#| error: true
old$date - old$date[1]
```

`non-numeric argument to binary operator` is R saying you asked it to subtract text from text. This is why every base-R set of instructions you will ever read has a line like `df$date <- as.Date(df$date)` near the top, and why you do not need one. `read_csv()` already did it.

There are two smaller differences. `read.csv()` returns `int` for whole-number columns where `read_csv()` returns `dbl`, and before R 4.0.0 it turned character columns into factors, which is the historical accident described in @sec-tibbles. It is also slower on large files, though at 345 rows you will never notice.

`read.csv()` is not broken and you may use it. The reason this book does not is that it tells you nothing on the way in.

## Pinning the types {#sec-col-types}

Guessing is right for exploring a new file and wrong for a script that runs every month. readr guesses from the first 1,000 rows by default, so a column that is numeric for a thousand rows and contains the word `pending` on row 1,500 will be guessed wrong. The following month's file changes, the guess changes with it, and your code produces different numbers without saying anything.

The fix is to state the types. `cols()` lets you name the ones you care about and set a fallback for the rest:

```{r}
#| label: col-types
covid_pinned <- read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols(
    administrative_area_level_1 = col_character(),
    date                        = col_date(format = "%Y-%m-%d"),
    .default                    = col_double()
  )
)
dim(covid_pinned)
```

No specification message this time, because there was nothing to guess. The quickest way to write one of these is to read the file once, run `spec()`, and paste the output into your script as the `col_types` argument. From then on, the file has to match your expectations or you hear about it.

Hearing about it looks like this. Insist that every column is numeric, including the two that are not:

```{r}
#| label: bad-types
bad <- read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols(.default = col_double())
)
```

A warning rather than an error, and the data still loads, with `NA` wherever a value could not be parsed. `problems()` tells you exactly which cells failed:

```{r}
#| label: problems
head(problems(bad))
```

Row 2, column 1, expected a double, found `Singapore`. That is a report you can act on. Two columns times 345 rows is 690 failures, and if a real import produces a handful rather than hundreds, those handful are usually the interesting rows: the ones where somebody typed a note into a numeric field.

When you only need part of a wide file, `cols_only()` reads the named columns and skips the rest, which on a large file saves both time and memory:

```{r}
#| label: cols-only
read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols_only(
    date      = col_date(),
    confirmed = col_double(),
    deaths    = col_double()
  )
)
```

The columns come back in the order they appear in the file, not the order you listed them, which catches people who then index by position.

### Other formats

`read_tsv()` for tab-separated, `read_delim()` when you have to name the separator yourself, `readxl::read_excel()` for `.xlsx` files including a `sheet` argument, and the `haven` package for Stata, SPSS and SAS files, which is the one that matters most in public health. All of them return tibbles, and `haven` keeps the variable labels that survey datasets carry.

## Always look at what you loaded {#sec-glimpse}

`glimpse()` is the first thing to run after any import. It transposes the usual view: one line per column, running across the page, so a 13-column table fits on a screen.

```{r}
#| label: glimpse
glimpse(covid)
```

Read it as three answers at once. The dimensions confirm the import. The type in angle brackets after each name confirms the parse. And the first handful of values confirms the content, which is where the interesting news usually is.

Look down the first values here and seven of the thirteen columns open with `NA`. Two different things are going on, and they are missing for different lengths of time.

```{r}
#| label: na-counts
sum(is.na(covid$deaths))
sum(is.na(covid$parks_percent_change_from_baseline))
```

`deaths` is missing for the first 59 rows. The first value the file carries is 2, on 21 March 2020, and it goes straight to 2 without ever passing through 1. Everything before that is stored as missing rather than as zero. Those are different claims. Zero says nobody died; `NA` says the source is not telling you. The distinction will cost you two deaths later in the book, when you try to recover daily deaths by differencing the column.

The six mobility columns are missing for the first 24 rows, up to and including 14 February. Google's Community Mobility Reports express each day as a percentage change from a baseline computed over 3 January to 6 February 2020, so the series cannot start until that baseline period has finished. The chapter on missing data takes both of these apart properly; for now, the point is that `glimpse()` showed you within seconds of loading the file.

Three more worth having in your fingers. `head()` and `tail()` show the first and last rows, and `tail()` is the one that catches the summary row somebody left at the bottom of a spreadsheet. `summary()` gives a five-number summary of every numeric column plus a count of `NA`s. `View()` opens the spreadsheet viewer in RStudio, which is the fastest way to eyeball a small file and useless on a large one; keep it at the console and never in a script, because it does nothing in a rendered document.

```{r}
#| label: view
#| eval: false
View(covid)
```

## The helper this book uses {#sec-read-covid}

Now `read_covid()` should hold no mystery. It calls `read_csv()` on this file via `here()`, with the column-types message switched off because you have just read it:

```{r}
#| label: helper
#| eval: false
read_covid <- function() {
  readr::read_csv(here::here("data", "sg_covid_data.csv"), show_col_types = FALSE)
}
```

Nothing is cleaned, nothing is repaired. Counts are still cumulative, `NA`s are still `NA`s. Every chapter that fixes something fixes it where you can watch.

## One plot, as a promise {#sec-first-plot}

You have not been taught `ggplot()` yet. Run this anyway.

```{r}
#| label: reward-plot
covid <- read_covid()

ggplot(covid, aes(x = date, y = confirmed)) +
  annotate(
    "rect",
    xmin = as.Date("2020-04-07"), xmax = as.Date("2020-06-01"),
    ymin = -Inf, ymax = Inf,
    fill = nus_colours[["grey"]], alpha = 0.4
  ) +
  geom_line(linewidth = 0.8) +
  annotate(
    "text",
    x = as.Date("2020-05-04"), y = 50000,
    label = "circuit breaker", size = 3.2, colour = nus_colours[["muted"]]
  ) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(
    title = "Singapore's 2020 epidemic, exactly as the file records it",
    subtitle = "Cumulative confirmed cases. Every point is a running total, not a daily count.",
    x = NULL,
    y = "Confirmed cases",
    caption = "Source: sg_covid_data.csv, 22 January to 31 December 2020."
  )
```

Fifty-eight thousand confirmed cases by the end of December, and a curve that turns sharply upward in April and flattens through the second half of the year. The shaded band is Singapore's circuit breaker, 7 April to 1 June 2020.

The curve is honest and almost useless for reading an epidemic, because a cumulative total can only go up. You cannot see when transmission accelerated or when it stopped. To see that you need the daily difference, which means creating a column, which is the next chapter.

## Exercise

::: {#exr-first-look}
## Land the file yourself

Start a fresh R session, load the tidyverse and `here`, and do this without using `read_covid()`.

1. Read `data/sg_covid_data.csv` with `read_csv()` and a path built by `here()`. Write down, in words, the type readr assigned to `date`, to `confirmed` and to `administrative_area_level_1`, and where in the message you found each one.
2. Read it a second time with the three types above stated explicitly and everything else left as double, and confirm the specification message no longer appears.
3. Run `glimpse()` and count the columns whose first value is `NA`. Then find the number of missing values in each of those columns, and say whether they all have the same number.
4. Read the same file with `read.csv()` instead. Find the one difference in the `date` column that would have forced you to write extra code, and demonstrate it failing.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

For part 2, `cols()` takes named columns plus a `.default` for everything you did not name. For part 3, `is.na()` returns a logical vector and `sum()` counts the `TRUE`s. For part 4, `class()` on the column is the fastest tell, and the failure you want is any arithmetic on a date.
:::

[Worked solution](solutions.qmd#sec-sol-first-look)
