# Welcome {.unnumbered}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

This is a book about getting an answer out of a dataset, written for people who have never programmed. It began as the week-one worksheet for SPH5421 at the NUS Saw Swee Hock School of Public Health: a few pages of base R with gaps to fill in, no answer key, and, I should own up to this, code that had never been run from top to bottom. Students would hit an error on question four and have no way of knowing whether the error was theirs or mine.

Everything you see here has been run. The outputs on these pages, the plots, the warnings, the place where R prints `NaN` instead of a number, are all the real thing, produced by the code shown directly above them. If a number in the prose disagrees with the number in the output block, the prose is wrong and I would like to know about it.

## Who this is for

Public health masters students, mostly. If you have used Stata or SPSS through a menu, or done your analysis in Excel and felt the spreadsheet slipping out from under you around row 4,000, you are the reader I had in mind. No programming background is assumed. I do assume you have seen a mean and a median before, and that you know roughly what an epidemic curve is.

By the time you reach the end you should be able to read a CSV into R and form an opinion about whether to trust it; turn cumulative counts into daily ones without silently losing the first day; find missing values, understand why they are missing, and decide what to do rather than letting R decide for you; group a dataset by month and summarise it; build a figure you would be willing to put in front of a supervisor; and write a small function so that the same calculation done six times is written once. The last chapter asks you to produce a monthly situation report from the raw file, with no scaffolding at all.

You will also, and this matters more than the syntax, learn to read an error message.

## The data

One dataset, all the way through. It is Singapore's 2020 COVID-19 record: 345 daily rows running from 22 January to 31 December 2020, with 13 columns. Cumulative confirmed cases and deaths, the three Oxford government-response indices, and the six Google Community Mobility series. It ends with 58,599 confirmed cases and 29 recorded deaths.

The file was assembled from the [COVID-19 Data Hub](https://covid19datahub.io) (Guidotti & Ardia, 2020), which in turn draws on Singapore's Ministry of Health, the Oxford COVID-19 Government Response Tracker, and Google's Community Mobility Reports. Appendix B documents every column, including where it is missing and why.

I chose it because it is small enough to look at whole and broken in all the ordinary ways. The death column has 59 leading missing values, and when it finally starts it starts at 2 rather than at 1, so the obvious way of turning it into daily deaths recovers 27 of the 29. Every mobility column is missing its first 24 days, because Google's baseline period had not finished when the series begins. The case counts are cumulative, so a plot of the raw column is a line that only ever goes up, and it tells you almost nothing about when things were bad. And there is a day in August with 908 cases sitting in a month whose median is 91, which is a reporting artefact rather than a wave. Every one of those turns up later as a lesson.

Here is where the book is going. The code below is the sort of thing you will be writing by the end of Part IV, and none of it needs to make sense yet.

```{r}
#| label: welcome-curve
#| fig-cap: "New confirmed cases per day, computed from the cumulative column in the raw file. The single tall bar in August is the 908-case reporting day, not a second wave."
covid <- read_covid()

covid %>%
  arrange(date) %>%
  mutate(daily_cases = confirmed - lag(confirmed, default = 0)) %>%
  ggplot(aes(x = date, y = daily_cases)) +
  geom_col() +
  labs(
    title = "The shape of Singapore's 2020",
    subtitle = "Peak of 1,426 new cases on 20 April, during the circuit breaker",
    x = NULL,
    y = "New cases"
  )
```

`read_covid()` is a shortcut defined in the book's setup file so that every chapter starts from the same place. The import chapter unwraps it and shows you the `read_csv()` call inside.

## Exercises, hints and solutions

Exercises sit inside the chapters, near the material they exercise. They state a goal in words. They do not give you a code skeleton with holes in it, because filling in holes teaches you to fill in holes.

Under most exercises there is a collapsed hint. Open it when you are stuck in the sense of not knowing which function to reach for. Under that, a link to the worked solution in Appendix A. The solutions are complete and commented, and they are deliberately one click away rather than upside down at the bottom of the page. You are an adult. If you want the answer you can have it. You will learn more if you have made a real attempt first, and you will know perfectly well whether you have.

::: {.callout-warning}
## SPH5421 students: where to stop

For the live session, work Parts I to IV: everything up to and including *Showing it to someone*. That is the material we will use in the rest of the module, and it is about as much as anyone absorbs in one sitting.

The whole book is roughly five and a half hours if you type the code rather than skim it. Parts V and VI, on writing your own functions and the capstone, are there for the week after, or for when you meet a problem that needs them. Nobody expects you to arrive having finished the book.
:::

## How to use this book

Type the code. Do not paste it. I know that sounds like an instruction from someone who enjoys watching people suffer, so here is the reasoning: pasting moves characters from one window to another and teaches your hands nothing. Typing makes you notice that `aes()` is inside `ggplot()` and not next to it, and it makes you produce your own typos in a safe setting where the fix is on the screen above. The error you generate from your own missing bracket in week one is the error you will recognise instantly at 11pm in month six. There is a copy button on every code block on this site. Use it when you are re-running something for the fifth time, not the first.

Read error messages to the end. Beginners tend to see red text, assume it is undifferentiated punishment, and start changing things at random. R's errors are usually two or three lines and the useful line is rarely the first. The tidyverse ones in particular put the diagnosis under a bullet several lines down, along with a suggestion for what to do about it. When you search for a message, paste it in exactly as R printed it, but delete your own object names first, because nobody else on the internet has a variable called `apr_data2`.

Run everything as you read. When a chunk produces something, look at it before scrolling on. Then change one thing and run it again. Change the `100` to `50`, swap `mean` for `median`, remove the `na.rm = TRUE` and see what breaks. This book is arranged so that breaking it is cheap.

Chapters render in separate R sessions and each one is self-contained, so you can restart R whenever you like and pick up at the top of a chapter. That is a habit worth having anyway, and the good-practice chapter explains why.

## Corrections

Every page has an "Edit this page" link that goes to its source on GitHub, and there is an issue tracker. A wrong number in a teaching resource propagates into other people's work, so corrections are welcome and I would rather hear about one than not.

## Acknowledgement

Written by Swapnil Mishra, working with Claude Code (Anthropic, Opus 5 model), which helped draft the text, convert the original base-R worksheet to the tidyverse, and check the numbers against the data. Responsibility for the result is entirely human.
