# Welcome {.unnumbered}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

This book covers getting an answer out of a dataset, and it is written for people who have never programmed. It began as the week-one worksheet for SPH5421 at the NUS Saw Swee Hock School of Public Health: a few pages of base R with gaps to fill in, no answer key, and code that had never been run from top to bottom. Students would hit an error on question four with no way of knowing whether the error was theirs or mine.

Everything here has been run. The outputs on these pages, including the plots, the warnings and the place where R prints `NaN` instead of a number, were produced by the code shown directly above them. If a number in the prose disagrees with the number in the output block, the prose is wrong and I would like to know about it.

## Who this is for

Public health masters students, mostly. The intended reader has used Stata or SPSS through a menu, or has done their analysis in a spreadsheet and found it difficult to keep track of at a few thousand rows. No programming background is assumed. I do assume you have seen a mean and a median before, and that you know roughly what an epidemic curve is.

By the time you reach the end you should be able to read a CSV into R and form a view about whether to trust it; turn cumulative counts into daily ones without silently losing the first day; find missing values, work out why they are missing, and decide what to do rather than letting R decide for you; group a dataset by month and summarise it; build a figure suitable for a report; and write a small function so that the same calculation done six times is written once. The last chapter asks you to produce a monthly situation report from the raw file, with no scaffolding.

The book also spends time on reading error messages, which is a skill separate from the syntax and is treated as such throughout.

## The data

One dataset, all the way through. It is Singapore's 2020 COVID-19 record: 345 daily rows running from 22 January to 31 December 2020, with 13 columns. Cumulative confirmed cases and deaths, the three Oxford government-response indices, and the six Google Community Mobility series. It ends with 58,599 confirmed cases and 29 recorded deaths.

The file was assembled from the [COVID-19 Data Hub](https://covid19datahub.io) (Guidotti & Ardia, 2020), which in turn draws on Singapore's Ministry of Health, the Oxford COVID-19 Government Response Tracker, and Google's Community Mobility Reports. Appendix B documents every column, including where it is missing and why.

I chose it because it is small enough to inspect in full and because it has the ordinary defects. The death column has 59 leading missing values, and when it does start it starts at 2 rather than at 1, so the obvious way of turning it into daily deaths recovers 27 of the 29. Every mobility column is missing its first 24 days, because Google's baseline period had not finished when the series begins. The case counts are cumulative, so a plot of the raw column rises monotonically and shows little about when incidence was high. And there is a day in August with 908 cases in a month whose median is 91, which is a reporting artefact rather than a wave. Each of these appears later as a worked example.

The code below is the sort of thing you will be writing by the end of Part IV. None of it needs to make sense yet.

```{r}
#| label: welcome-curve
#| fig-cap: "New confirmed cases per day, computed from the cumulative column in the raw file. The single tall bar in August is the 908-case reporting day, not a second wave."
covid <- read_covid()

covid %>%
  arrange(date) %>%
  mutate(daily_cases = confirmed - lag(confirmed, default = 0)) %>%
  ggplot(aes(x = date, y = daily_cases)) +
  geom_col() +
  labs(
    title = "The shape of Singapore's 2020",
    subtitle = "Peak of 1,426 new cases on 20 April, during the circuit breaker",
    x = NULL,
    y = "New cases"
  )
```

`read_covid()` is a shortcut defined in the book's setup file so that every chapter starts from the same place. The import chapter unwraps it and shows you the `read_csv()` call inside.

## Exercises, hints and solutions

Exercises sit inside the chapters, near the material they exercise. Each states a goal in words and leaves the structure of the answer to you, so deciding how to build the answer is part of the work.

Under most exercises there is a collapsed hint, which names the function or approach to reach for. Under that is a link to the worked solution in Appendix A. The solutions are complete and commented, and each is one click away. Attempting the exercise first is more useful than reading the solution first, but the choice is yours.

::: {.callout-warning}
## SPH5421 students: where to stop

For the live session, work Parts I to IV: everything up to and including *Showing it to someone*. That is the material the rest of the module builds on.

The whole book takes roughly five and a half hours if you type the code rather than skim it. Parts V and VI, on writing your own functions and the capstone, are for the week after, or for when you meet a problem that needs them. You are not expected to arrive having finished the book.
:::

## How to use this book

Type the code rather than pasting it. Typing makes the structure visible: that `aes()` sits inside `ggplot()` rather than next to it, for instance. It also produces your own typos in a setting where the correct version is on the screen above, so you meet the common error messages while their cause is still obvious. There is a copy button on every code block on this site, which is useful when you are re-running something for the fifth time.

Read error messages to the end. R's errors usually run to two or three lines and the useful line is rarely the first. The tidyverse ones put the diagnosis under a bullet several lines down, along with a suggestion for what to do about it. When you search for a message, paste it in exactly as R printed it, but delete your own object and file names first, since those appear nowhere outside your own session.

Run everything as you read. When a chunk produces something, look at it before scrolling on, then change one thing and run it again. Change the `100` to `50`, swap `mean` for `median`, remove the `na.rm = TRUE` and see what changes. Every chunk can be re-run from the top of its chapter, so an experiment that fails costs nothing.

Chapters render in separate R sessions and each one is self-contained, so you can restart R at any point and pick up at the top of a chapter. The good-practice chapter sets out why restarting frequently is worth doing.

## Corrections

Every page has an "Edit this page" link that goes to its source on GitHub, and there is an issue tracker. A wrong number in a teaching resource propagates into every analysis built on it, so corrections are welcome, however small.

## Acknowledgement

Written by Swapnil Mishra, working with Claude Code (Anthropic, Opus 5 model), which helped draft the text, convert the original base-R worksheet to the tidyverse, and check the numbers against the data. Responsibility for the result is entirely human.
