# Where to go next {#sec-next-steps}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

```{r}
#| label: data
#| include: false
covid <- covid_daily()
```

You now know enough R to be dangerous, which is the honest description of where a week gets you.
You can read a file, check it, reshape it, summarise it by group, plot it, and write a function
when the same six lines show up for the third time. What you don't yet have is the collection of
small recoveries that turn an error message from a wall into an inconvenience. That comes from
hours, not from chapters.

This last one is about where those hours are best spent.

## What the rest of the course does with all this

The modelling weeks turn the direction of travel around. Everything so far started with data and
went looking for a description. A compartmental model starts with a description, some
assumptions about how people move between states, and produces data you can compare with the
real thing.

The mechanics are less exotic than they sound. You divide a population into compartments,
susceptible, infectious, recovered in the simplest case. You write down the rate at which people
leave each compartment and enter the next. Then you hand those rates to a solver, which steps
them forward in small increments and hands back a trajectory.

In R the solver is usually [deSolve](https://cran.r-project.org/package=deSolve), and its main
function is `ode()`. The shape of the code is this.

```{r}
#| label: sir-sketch
#| eval: false
sir <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS <- -beta * S * I / N
    dI <-  beta * S * I / N - gamma * I
    dR <-  gamma * I
    list(c(dS, dI, dR))
  })
}

state      <- c(S = 99999, I = 1, R = 0)
parameters <- c(beta = 0.4, gamma = 0.2, N = 1e5)
times      <- seq(0, 180, by = 1)

out <- ode(y = state, times = times, func = sir, parms = parameters)
```

Read that against what you already know and almost all of it is familiar. `sir` is a function
with three arguments whose value is whatever the last expression evaluates to, which is the
functions chapter. `state` and `parameters` are named numeric vectors, and `as.list()` and
`with()` are there so you can write `beta` instead of `parameters[["beta"]]`, which is the
foundations chapter. The return value is a list containing one vector, because `ode()` insists
on a list, and knowing the difference between `c(dS, dI, dR)` and `list(c(dS, dI, dR))` is why
that chapter spent time on lists.

What `ode()` gives back is a matrix, not a tibble. Your first move will be `as_tibble()`, then
`pivot_longer()` so you can draw three lines on one plot, then the ggplot2 grammar from the
visualising chapter. Two evenings of this course will be spent on models and about ten minutes
of each of them on plotting, because you already did the plotting.

Two things will catch you, so they're worth naming now.

The `time` column that comes out of `ode()` counts steps from zero. It is not a date. If you
want to line a trajectory up against the `daily_cases` column you built, you have to decide what
day zero means and add it: `start_date + out$time`. Model time and calendar time are different
quantities and mixing them silently is a classic way to get a plot that looks fine and is wrong
by a fortnight.

The second is that you will want to run the model more than once. Sweeping over ten values of
beta, or a hundred, is the iteration chapter's problem exactly: write the model run as a function
of the parameter you're varying, then map over the values and bind the results, rather than
copying the call ten times with a number changed.

Once your models get big enough that speed matters, or you're running thousands of stochastic
replicates, R's own loop overhead starts to hurt. [odin](https://cran.r-project.org/package=odin)
is the usual answer: you write the equations in a small R-like language and it generates and
compiles C for you. The equations look much like the ones above. The habits of writing them
carefully, naming parameters properly and checking a trajectory before believing it are the same
habits.

Fitting a model to Singapore's 2020 curve is where the two halves of the course meet, and it's
also where the caveats from the capstone come back. You'd be fitting to reported cases, which are
a function of testing as well as transmission, over a period when the national count was
dominated by the dormitory outbreak. A model fitted to that series without a term for either is
answering a question you didn't ask.

## Five things worth reading

Short list, because a long one gets ignored.

[**R for Data Science**, 2nd edition](https://r4ds.hadley.nz/), by Wickham, Çetinkaya-Rundel and
Grolemund. Free online. This is the book this course is a week-long trailer for. The chapters on
data transformation and visualisation cover the same ground with more room, and the chapters on
joins, factors, strings and dates cover the things I had to leave out. If you read one thing
from this list, read this one, and read it with your hands on the keyboard.

[**The Epidemiologist R Handbook**](https://epirhandbook.com/en/), by Neale Batra and a large
group of contributors. Free online. This is a recipe book rather than a textbook, written by
field epidemiologists for field epidemiologists, and it is the fastest route from "I need an epi
curve with a moving average and a case fatality ratio by age group" to working code. Its examples
are outbreak data. Use it when you know what you want and not how to spell it.

[**Happy Git and GitHub for the useR**](https://happygitwithr.com/), by Jenny Bryan. Free online.
Version control is the single habit with the best ratio of effort to regret avoided, and this is
the book that gets R users over the initial hostility of git. Read the first eight or so chapters,
get one project under version control, stop there. You do not need to understand rebasing.

[**The tidyverse style guide**](https://style.tidyverse.org/), by Hadley Wickham. An afternoon's
read, mostly skimmable. It's where the conventions this book uses come from: spaces around `<-`,
lowercase names with underscores, one dplyr verb per line. Consistency matters less for its own
sake than because it lets you read your own code from three months ago without decoding it first.

[**Advanced R**, 2nd edition](https://adv-r.hadley.nz/), by Hadley Wickham. Free online. Not for
now. Bookmark it and come back when you find yourself asking why something behaves the way it
does rather than how to make it work. The chapters on environments and on functionals are the
ones that explain the odd corners of R that this book has been telling you to accept.

Notice what is not on that list. There's no reference manual, because R's own help pages are the
reference manual and `?filter` is faster than any book. There's no book on statistics in R,
because the statistics course is a statistics course and the R is the easy part of it.

## When it breaks

Something will break tonight. Here is the order to do things in.

### Read the message, all of it

R's error messages have improved enormously and most people still read the first line and panic.
The tidyverse ones in particular are built to be read from top to bottom. Here's one you can
produce on purpose.

```{r}
#| label: broken
#| error: true
month_summary <- function(data, month_start) {
  data %>%
    filter(month == month_start) %>%
    summarise(cases = sum(daily_cases), peak = max(daily_cases))
}

month_summary(covid, "2020-08")
```

Four lines, four separate pieces of information. The first says which verb failed: `filter()`,
not `summarise()`, so the pipeline died before it got to the summary. The reference to
`dplyr/R/summarise.R` is dplyr telling you where in its own source the call was recorded, and it
is not about your file. The second line names the argument that blew up, `month == month_start`.
The third says the actual failure happened inside `charToDate()`, which is a base R function for
turning text into dates and not a function you called. The fourth is the complaint itself.

Put those together and the story is complete. Comparing a `Date` column with a character string
makes R try to convert the string to a date, and `"2020-08"` isn't a date because R has no idea
whether you meant the first of the month or the whole of it. It refuses to guess.

```{r}
#| label: fixed
month_summary(covid, as.Date("2020-08-01"))
```

The habit worth building: when you read an error, work out which line names *your* code and which
lines name the machinery underneath it. The fix is almost always at the boundary between them.

If you're working in RStudio you'll also see `Run rlang::last_trace() to see where the error
occurred.` Do that. It prints the full chain of calls with your own function at the top and the
failing one at the bottom, and for a pipeline buried inside a function it's often the quickest
way to find which of your six lines actually broke.

### traceback(), for the errors that don't tell you where they came from

Base R errors are terser. This is one from a function calling another function.

``` r
> daily_mean <- function(x, i, k) mean(x[(i - k):(i + k)])
> smooth_at <- function(data, i) daily_mean(data$daily_cases, i, 3)
> smooth_at(covid, "100")
Error in i - k : non-numeric argument to binary operator
```

That tells you what went wrong and says nothing about where. `traceback()`, run immediately
afterwards, does.

``` r
> traceback()
3: mean(x[(i - k):(i + k)]) at #1
2: daily_mean(data$daily_cases, i, 3) at #1
1: smooth_at(covid, "100")
```

Read it from the bottom. Line 1 is the call you typed. Line 2 is what it called. Line 3 is where
it died. So `smooth_at()` passed the string `"100"` straight through to `daily_mean()`, which
tried to subtract 3 from it. The `at #1` is the line number in the source; when your functions
live in a file rather than being typed at the console, you get something like `at report.R#12`,
which is the number you want.

RStudio puts a "Show Traceback" button next to errors in the console. Same thing, one click.

### browser(), for when you need to be inside it

When the traceback tells you which function died but not why, put `browser()` on the line before
the trouble and run the call again. R stops there and hands you a prompt with the function's own
variables in scope.

```{r}
#| label: browser-demo
#| eval: false
month_summary <- function(data, month_start) {
  browser()
  data %>%
    filter(month == month_start) %>%
    summarise(cases = sum(daily_cases))
}

month_summary(covid, "2020-08")
```

``` r
Called from: month_summary(covid, "2020-08")
Browse[1]> month_start
[1] "2020-08"
Browse[1]> class(month_start)
[1] "character"
Browse[1]> class(data$month)
[1] "Date"
Browse[1]> Q
```

Three questions and the bug is obvious: you're comparing a character against a Date. Type `Q` to
quit the browser, `n` to run the next line, `c` to let it finish. Then take the `browser()` back
out, which everyone forgets to do at least once and then wonders why their script keeps stopping.

`debugonce(month_summary)` does the same thing without editing the function, which is the better
move when the function belongs to a package.

### The failures that don't announce themselves

Errors are the good case. R stops and tells you. The dangerous failures are the ones that return
a number.

```{r}
#| label: silent
covid$date[which.max(covid$daily_case)]
```

One missing letter in the column name. R gives you a warning and a zero-length date rather than
an error, because `$` on a tibble returns `NULL` for a column that isn't there and `which.max()`
of nothing is nothing. Inside a longer pipeline that warning scrolls past and you publish a
report with a blank where the peak date should be.

Here's a worse one, and it's from this dataset.

```{r}
#| label: deaths-gap
max(covid$deaths, na.rm = TRUE)
sum(covid$daily_deaths, na.rm = TRUE)
```

Singapore recorded 29 COVID-19 deaths in 2020. Differencing the cumulative column recovers 27 of
them, because `deaths` has 59 leading missing values and the first number that appears is already
2, on 21 March. The two missing deaths are hidden inside that first observation. Neither line
above errors. Neither warns. One of them is wrong for the question "how many people died", and
which one depends entirely on what you meant.

The defence isn't cleverness, it's arithmetic you can check. Does the total match a number you
know from somewhere else? Does the number of rows match the number of days? Does the maximum land
on the date you expected? Every one of those takes a line of code and catches the kind of mistake
that no amount of careful reading will.

## How to ask a question that gets answered

You will get stuck in a way that fifteen minutes of your own effort won't fix. When that happens,
how you ask determines whether you wait five minutes or five days.

A question that gets answered has four parts. What you were trying to do, stated in terms of the
result you wanted. The code you ran, as text. What actually happened, with the error message
copied exactly. And what you already tried, so nobody wastes their time suggesting it.

The exact error text matters more than anything else in that list. Paste it, don't describe it.
"It says something about a non-numeric argument" is unsearchable. `non-numeric argument to binary
operator` is one search away from a thousand people who've had the same problem. Before you ask
anyone, put the message in a search engine with the word `r` and the name of the package. That
solves a good half of everything.

Make it small. If your pipeline is forty lines, cut it in half and see if the error survives.
Keep cutting. Most bugs collapse to three or four lines, and the act of cutting finds the answer
often enough that it's worth doing even when you're planning to ask anyway. If the problem needs
data, don't attach a 30 MB CSV. `dput(head(covid, 5))` prints a few rows as code that anyone can
paste into their own session.

The [reprex](https://cran.r-project.org/package=reprex) package automates the tidy version of
this. You copy your minimal example to the clipboard, run `reprex::reprex()`, and it runs the
code in a clean session and puts a formatted block on your clipboard with the code and its real
output interleaved, ready to paste.

```{r}
#| label: reprex-demo
#| eval: false
reprex::reprex({
  library(dplyr)
  x <- c(1, 2, NA)
  mean(x)
})
```

The clean session is the point. It catches the case where your code only works because of
something you loaded an hour ago and forgot about, which is a surprisingly large fraction of
problems that feel like package bugs.

Include your R version and package versions when the problem smells like a version difference.
`sessionInfo()` prints them.

Ask your classmates first. They're on the same data, the same versions and the same week, and
somebody hit your problem yesterday. After that, the
[Posit Community](https://forum.posit.co/) forum is patient with beginners, and Stack Overflow is
faster but less forgiving, so read a few answered R questions there before posting your first one.
Report an issue on a package's GitHub only once you can show the maintainer a reproducible example
and you're fairly sure the package is at fault rather than you.

One thing not to do: don't post a photograph of your screen. Nobody can copy code out of a
photograph, and the people most able to help you are the ones least willing to retype it.

And when you work something out that took you two hours, write down what the fix was, in a
comment in the script, in the words you'd have needed at the start. You'll meet it again in
March. So will whoever asks the question after you.
