# Rectangles: data frames and tibbles {#sec-rectangles}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

The worksheet this book replaces asked, in prose, why data frames are useful, and left three ruled lines for the answer. Nobody ever filled them in, and the model answer was never written. So here is the answer, and it is not a matter of convenience. The data frame is the reason R can do epidemiology at all.

```{r}
#| label: load
covid <- read_covid()
```

## Three vectors and a problem {#sec-loose-vectors}

Take three days of the Singapore record, in the week the circuit breaker began, and store each variable in its own vector.

```{r}
#| label: loose
day        <- as.Date(c("2020-04-06", "2020-04-07", "2020-04-08"))
confirmed  <- c(1375, 1481, 1623)
stringency <- c(45.37, 62.04, 76.85)
```

Everything you need is there. The correspondence between them is not written down anywhere; it lives entirely in your head, as a promise that the second element of each vector refers to the same day. R has made no such promise, and will not enforce one.

Watch it break. Sort the stringency values from highest to lowest, which is a reasonable thing to want:

```{r}
#| label: sort-breaks
sort(stringency, decreasing = TRUE)
day
```

`stringency` has been reordered. `day` has not. If you assign that sorted vector back over the old one, the 7 April lockdown score is now sitting next to 6 April, and nothing in your session knows. No error, no warning, and a plot that looks entirely plausible. This is the kind of mistake that survives peer review.

Multiply the problem by 345 rows and 13 variables and you have the reason data frames exist.

## A data frame is a list of columns {#sec-list-of-columns}

`data.frame()` binds vectors together so that they can only be manipulated as a unit.

```{r}
#| label: build-df
three_days <- data.frame(day, confirmed, stringency)
three_days
```

Now sorting means sorting *rows*, and the rows carry their contents with them:

```{r}
#| label: sort-rows
three_days[order(three_days$stringency, decreasing = TRUE), ]
```

The row numbers on the left have gone 3, 2, 1. Each row moved as a piece.

Underneath, a data frame really is a list of vectors, and you can see that by stripping the class off it:

```{r}
#| label: unclass
unclass(three_days)
```

A named list of three vectors, plus an attribute recording the row names. That is the whole implementation. Everything else, the printing, the `nrow()`, the two-dimensional subsetting, is behaviour attached to that list by its class. R will confirm it if you ask directly:

```{r}
#| label: is-list
is.list(three_days)
length(three_days)
lengths(three_days)
```

`length()` on a data frame counts the elements of the underlying list, which is the number of columns. `lengths()`, with the s, gives the length of each element, which is the number of rows repeated once per column. `nrow()` and `ncol()` are friendlier names for the same two facts.

Two properties follow directly from "list of vectors", and they are the properties that matter.

**Columns may be of different types.** A list can hold anything, so one element can be a `Date` vector, the next a `double` vector, the next a `character` vector. Each column is internally uniform, which is the vector rule from @sec-types, but columns need not agree with each other. That is what makes a data frame the right shape for a dataset, where dates, counts and labels sit side by side. A matrix cannot do this; every cell of a matrix has one type.

```{r}
#| label: col-types
sapply(three_days, class)
```

**Rows must line up.** The list is not free to hold vectors of any length, because the rectangle would stop being a rectangle. Give `data.frame()` mismatched lengths and it refuses:

```{r}
#| label: mismatch
#| error: true
data.frame(day = 1:3, phase = c("pre", "circuit breaker"))
```

`arguments imply differing number of rows: 3, 2` is a good error. It arrives at the moment the mistake is made, not three steps later.

That constraint is the entire point. Every operation on a rectangle, filtering rows, creating columns, grouping and summarising, is built on the guarantee that row *i* of every column refers to the same observation. `filter()` can drop 200 rows without you checking that anything stayed aligned, because misalignment is not representable.

### Where `data.frame()` lets you down

The constraint has one hole in it, and it is old and unfortunate. If the short vector's length divides the long one exactly, `data.frame()` recycles instead of complaining:

```{r}
#| label: recycle-df
data.frame(day = 1:6, phase = c("pre", "circuit breaker"))
```

Six rows, alternating labels, no warning. That is the recycling rule from @sec-recycling applied where it does real damage. It is convenient when you intend it, and silent when you do not, which is a poor trade for data work.

## Tibbles {#sec-tibbles}

A tibble is a data frame with the historical annoyances removed. `read_csv()` returns one, so this is what you have been holding all along:

```{r}
#| label: class-covid
class(covid)
```

Read that from the right. It *is* a `data.frame`, so anything that accepts a data frame accepts it. `tbl` and `tbl_df` are the tibble classes, which change how it prints and how it is subsetted. `spec_tbl_df` is readr's marker, carrying the column specification it worked out while reading the file: which column it took to be a date, which it took to be numbers, which it left as text.

Three things changed, and each of them fixes something that used to cost people time.

### Printing tells you what you have

```{r}
#| label: print-tibble
covid
```

Ten rows, then a count of what was left out, then the names and types of the columns too wide to show. The `<chr>`, `<dbl>` and `<date>` labels under the column names are the part to pay attention to. Half the confusion in early R comes from a column being character when you assumed it was numeric, and a tibble tells you before you have written a line of analysis.

An old-style data frame does not do this. Print the same 345 rows as a `data.frame` and R sends all of them to the console, unlabelled, and you scroll for a while looking for the top. Try it once so you know what it looks like, then use `head()`:

```{r}
#| label: print-df
head(as.data.frame(covid)[, 1:5])
```

### No accidental factors

Before R 4.0.0, released in April 2020, `data.frame()` and `read.csv()` converted every character column into a factor by default, because the `stringsAsFactors` argument defaulted to `TRUE`. A factor stores its values as integer codes against a table of levels, which is what you want for a categorical variable in a model and emphatically not what you want for a column of free text or identifiers.

The behaviour is still available, and it still surprises people who ask for it:

```{r}
#| label: factors
phase <- data.frame(label = c("pre", "circuit breaker"), stringsAsFactors = TRUE)
phase$label
```

`Levels: circuit breaker pre` is alphabetical, so "circuit breaker" now sorts before "pre" and any plot you make will show them in that order. R 4.0.0 changed the default to `FALSE`, which is one of the few backwards-incompatible changes the language has ever made, and it was made because this had wasted enough people's time. Tibbles never did the conversion in the first place.

Factors are useful when you ask for one. A severity label that ought to read low, medium, high rather than alphabetically is exactly the case for `factor()`, because you name the levels in the order you want and everything downstream respects it. The difference is that it is your decision, written in code somebody can point at.

### Subsetting stays a data frame

Here is the change that saves the most debugging.

::: {.callout-important}
## Predict before you run

`old` below is the same data as `covid`, converted to an old-style data frame. Both of the next two expressions ask for one column with a single pair of square brackets. Decide, before running them, whether each returns a rectangle of 345 rows and 1 column, or a bare vector of 345 numbers.
:::

```{r}
#| label: drop-setup
old <- as.data.frame(covid)
class(old[, "confirmed"])
class(covid[, "confirmed"])
```

The data frame silently threw away a dimension and handed back a vector. The tibble kept the rectangle.

```{r}
#| label: drop-show
covid[, "confirmed"]
```

The old behaviour comes from `[.data.frame` having a `drop` argument that defaults to `TRUE`: ask for a single column and R assumes you wanted the contents rather than a one-column table. It is defensible in an interactive session and awful inside a pipeline, because a function two steps downstream that expected a data frame receives a numeric vector and fails with a message about the wrong thing entirely. Tibbles set `drop = FALSE` and never guess.

When you do want the bare vector, say so with `$` or double brackets, both of which mean "give me the contents of this element" and work identically on both kinds of rectangle:

```{r}
#| label: extract
class(covid$confirmed)
class(covid[["confirmed"]])
```

### No silent partial matching

`$` on an old-style data frame will complete a partial column name for you:

```{r}
#| label: partial-df
head(old$conf, 5)
```

You asked for `conf`. There is no column called `conf`. R found `confirmed`, decided that was probably what you meant, and returned it without comment. Now imagine a file that also has a `confirmed_cases_dormitory` column: the match becomes ambiguous and you get `NULL`, on a day when your code worked yesterday.

Tibbles warn:

```{r}
#| label: partial-tbl
head(covid$conf, 5)
```

A warning saying the column is unknown or uninitialised, then `NULL`. The same warning appears for a straightforward typo, which is how you find out that `covid$confrimed` is not a thing before it quietly poisons a calculation.

### The rule

Use tibbles for anything you write. When an older package rejects one, and a few statistical packages still do, convert at the point of the call:

```{r}
#| label: as-data-frame
class(as.data.frame(covid))
```

`as_tibble()` converts back. The two are compatible in both directions, so this is a small cost paid at one line, not a decision about your whole project.

## Tidy data {#sec-tidy}

Rectangles are the container. Tidy data is a rule about what goes in which part of it, and it has three clauses:

1. Each variable is a column.
2. Each observation is a row.
3. Each value is a cell.

The rule sounds obvious until you look at real spreadsheets, where it is broken constantly: twelve columns named January to December, a column holding both a measurement and its unit, two rows per patient because someone recorded before and after side by side.

Every tool in the tidyverse assumes tidy input. That is the whole bargain. Get the shape right once and `filter()`, `mutate()`, `group_by()` and `ggplot()` all work without argument.

So look at this file through that lens. What is one observation?

```{r}
#| label: unit
nrow(covid)
length(unique(covid$date))
length(unique(covid$administrative_area_level_1))
```

345 rows, 345 distinct dates, one country. The unit of observation is a country-day, and since the country never varies, in practice it is a day. Each column is a variable measured on that day: a cumulative case count, a cumulative death count, a policy index, six mobility series. One row per observation, one column per variable, one value per cell. The file arrives tidy, which is a piece of luck this book takes advantage of and your own data will not always grant you.

### A column that says nothing

`administrative_area_level_1` deserves a second look:

```{r}
#| label: constant-col
table(covid$administrative_area_level_1)
```

One distinct value across all 345 rows. As a variable it is dead: a column that never varies cannot explain, predict, group or distinguish anything. It is here because this extract was cut out of a multi-country archive where that column did vary, and slicing one country out left the label behind.

Counting distinct values in every column is a habit worth forming on any new dataset, because both ends of the result tell you something. A column with one distinct value carries no information. A column with as many distinct values as there are rows is an identifier rather than a measurement, which is what `date` is here. Anything in between is a variable you can actually work with.

```{r}
#| label: distinct-counts
sapply(covid, function(column) length(unique(column)))
```

`stringency_index` takes 13 distinct values over the year, which is a hint about how it is constructed: it is a score assembled from a handful of policy indicators, so it moves in steps when a policy changes rather than drifting continuously. Keep that in mind before you read meaning into every wobble in it.

### When the file is not tidy

You will be handed untidy files. The common shapes are a header row of dates or months, one column containing two variables mashed together, and observations split across several rows. `tidyr` has `pivot_longer()` and `pivot_wider()` for reshaping between them, and they are worth looking up on the day you need them. This book does not spend a chapter on reshaping, because the Singapore file does not need it, and learning to reshape data is much easier once the rest of this is second nature.

## Exercise

::: {#exr-tibble-anatomy}
## Take a tibble apart

Using `covid` as loaded at the top of this chapter:

1. Show, with code rather than assertion, that `covid` is a list, how many elements that list has, and that every element has the same length. Say in one sentence which of those two numbers is the number of rows and which is the number of columns.
2. Find out the type of every column without printing the whole tibble.
3. Build a small tibble by hand recording three dates in Singapore's 2020 policy timeline, 7 April, 2 June and 19 June, alongside a short label for each. Then explain in two sentences why holding those as a tibble is safer than holding the dates and the labels as two separate vectors.
4. Convert `covid` to an old-style data frame and find one behaviour, other than printing, that differs from the tibble. Show it.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

For part 1, three functions from this chapter do the whole job: one asks whether it is a list, one counts the elements, one measures each element. For part 2, `sapply()` applies a function to every column and collects the answers into one named vector.
:::

[Worked solution](solutions.qmd#sec-sol-tibble-anatomy)
