# Making decisions {#sec-conditionals}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

Much of routine epidemiological reporting turns numbers into labels: a count of 1,426 becomes a high day before it reaches a bulletin. R provides two different tools for this. One decides what the *script* does next; the other decides what goes in a *column*. Using the first where the second is needed produces either an error or, less visibly, a result that is correct for one row and wrong for the other 344.

This chapter distinguishes the two, then uses the second to build a severity label for every day of 2020.

## Where we are

Everything here starts from the daily counts you built in the previous chapter. The file ships cumulative totals, so a day's new cases is today's total minus yesterday's, and `lag(confirmed, default = 0)` supplies the "yesterday" for the very first row. That default matters: `confirmed` is 0 on 22 January, so day one had a real zero of new cases, not a missing one.

```{r}
covid <- read_covid() %>%
  arrange(date) %>%
  mutate(daily_cases = confirmed - lag(confirmed, default = 0))

covid %>%
  select(date, confirmed, daily_cases) %>%
  slice(85:90)
```

## `if` tests a single value

R's `if` is a control-flow statement. It looks at a single TRUE or FALSE and picks which block of code to run. It is the appropriate tool when the decision is about the analysis as a whole.

```{r}
worst <- max(covid$daily_cases)
worst

if (worst > 1000) {
  "Singapore recorded a four-figure day."
} else {
  "No day reached a thousand."
}
```

`worst` is one number, so `worst > 1000` is a single TRUE and `if` can act on it.

Suppose instead you want a label on every row, and write the version that reads like English:

```{r}
#| error: true
if (covid$daily_cases > 300) "High" else "Not high"
```

`covid$daily_cases > 300` is not a single TRUE or FALSE but a vector of 345 of them.

```{r}
length(covid$daily_cases > 300)
```

`if` has no way to act on 345 answers at once, so it refuses. The message to search for is **the condition has length > 1**.

Up to R 4.1 the same code ran: R issued a warning, took the first element and carried on. A script could label an entire dataset "Not high" because 22 January happened to have zero cases, with only a warning to indicate it. R 4.2 made this an error, so a script written against the earlier behaviour stops here on any current version of R.

When you do want a whole-dataset question, collapse the vector to one value first. `any()` and `all()` are built for this.

```{r}
if (any(covid$daily_cases > 1000)) {
  "At least one day topped a thousand."
}
```

## `if_else()` builds a column

For a label on every row you want a *vectorised* choice: one input vector in, one output vector out, same length. dplyr's `if_else()` does that.

```{r}
covid %>%
  mutate(big_day = if_else(daily_cases > 300, "High", "Not high")) %>%
  count(big_day)
```

Read the arguments as a sentence: for each row, test `daily_cases > 300`; where it is TRUE use `"High"`, where it is FALSE use `"Not high"`.

`if_else()` requires both branches to be of the same type, and the requirement is deliberate:

```{r}
#| error: true
covid %>%
  mutate(big_day = if_else(daily_cases > 300, "High", 0))
```

**Can't combine `true` <character> and `false` <double>** means the column would be text on some rows and a number on others. `if_else()` does not choose between them. Base R's `ifelse()` does choose, by coercing both branches to a common type, and the coercion drops attributes. The effect on a date column:

```{r}
big <- covid %>%
  filter(daily_cases > 300) %>%
  slice(1:3)

ifelse(big$daily_cases > 300, big$date, as.Date(NA))
```

Those five-digit numbers are days since 1 January 1970, which is how R stores a `Date` underneath. `ifelse()` stripped the class attribute and handed back the raw numbers, so a date column comes out as a numeric one.

```{r}
if_else(big$daily_cases > 300, big$date, as.Date(NA))
```

`if_else()` keeps the class of the branches it returns, so the same comparison comes back with the dates intact.

## `case_when()` when there are more than two answers

Each label beyond the second adds another test, and nesting `if_else()` inside `if_else()` produces code that is hard to read and harder to check. `case_when()` replaces the nesting with a list of conditions, each paired with the value to use when it matches.

The course uses three cutoffs: under 100 new cases is Low, 100 up to and including 300 is Medium, and over 300 is High.

```{r}
covid <- covid %>%
  mutate(
    # Course convention: fewer than 100 cases is Low, 100 up to and
    # including 300 is Medium, over 300 is High.
    severity = case_when(
      daily_cases <  100 ~ "Low",
      daily_cases <= 300 ~ "Medium",
      daily_cases >  300 ~ "High"
    )
  )

covid %>% count(severity)
```

The `~` separates the test from the result. R uses the tilde wherever the two sides of a relationship have to sit in one expression, which is why it also appears in modelling formulas and in ggplot facet specifications; inside `case_when()` it means "gives".

`case_when()` works top to bottom and **first match wins**. Once a row matches a line, later lines never see it. That is what lets the second condition be `daily_cases <= 300` and not `daily_cases >= 100 & daily_cases <= 300`: any row still being tested on line two already failed line one, so it already has at least 100 cases.

The order of the lines therefore matters. Swapping the first two leaves the arithmetic identical and the answer different:

```{r}
covid %>%
  mutate(
    wrong = case_when(
      daily_cases <= 300 ~ "Medium",
      daily_cases <  100 ~ "Low",
      daily_cases >  300 ~ "High"
    )
  ) %>%
  count(wrong)
```

Because the Medium line is evaluated first, it matches every Low day as well, and 215 days are mislabelled with no error and no warning. When a `case_when()` returns a category with a count of zero, or one much larger than expected, the order of the conditions is the first thing to check.

The other failure mode is a row that matches nothing at all. Drop the last line and the 77 High days match no condition:

```{r}
covid %>%
  mutate(
    incomplete = case_when(
      daily_cases <  100 ~ "Low",
      daily_cases <= 300 ~ "Medium"
    )
  ) %>%
  count(incomplete)
```

`case_when()` returns `NA` for unmatched rows rather than complaining, so this only shows up if you look. `.default` catches everything that falls through and is worth using whenever your conditions are meant to be exhaustive:

```{r}
covid %>%
  mutate(
    severity_alt = case_when(
      daily_cases <  100 ~ "Low",
      daily_cases <= 300 ~ "Medium",
      .default = "High"
    )
  ) %>%
  count(severity_alt)
```

Two details of that block are worth adopting. The recoded column gets a new name, `severity_alt`, rather than overwriting `severity`, so both versions remain available to count side by side while you check the change. The category labels are also spelled exactly as the existing column has them, because later steps match text literally: a lower-case `"high"` would start a fourth category rather than match `"High"`.

## Boundary values

"100 to 300" is ambiguous in English and unambiguous in code, so writing the code forces a decision that the specification may have left open: a day with exactly 100 cases could be Low or Medium, and 300 could be Medium or High. The `case_when()` settles both questions one way or the other, so the convention is worth choosing deliberately.

This dataset contains days sitting on both ends of the rule.

::: {.callout-important}
## Predict before you run

Consider two days in August: 18 August had exactly 100 new cases, and 6 August had 301. Under the cutoffs written above, write down the label each of them gets before you look. The boundary cases are the ones worth checking, because they are where an off-by-one in the cutoffs shows up.
:::

```{r}
covid %>%
  filter(date %in% as.Date(c("2020-08-06", "2020-08-18"))) %>%
  select(date, confirmed, daily_cases, severity)
```

18 August is Medium because `daily_cases < 100` is FALSE at exactly 100, so the day falls through to the `<= 300` line; 6 August is High because 301 is one case past `<= 300`. Changing `<` to `<=` on the first line moves 18 August to Low, so a single character changes the label on a day that occurs in the data.

Write the thresholds down in prose next to the code, with the inequalities spelled out, as the comment in the `case_when()` block above does. "Fewer than 100" and "up to 100" are different rules, and they disagree on days that occur in this dataset.

## Missing values fall straight through

Neither `if_else()` nor `case_when()` treats `NA` as FALSE. `NA > 0` is `NA`, which is neither branch, so the result is `NA`. Deaths in this file are missing for the first 59 days, so they show it plainly:

```{r}
covid %>%
  mutate(
    daily_deaths = deaths - lag(deaths, default = 0),
    death_day = if_else(daily_deaths > 0, "yes", "no")
  ) %>%
  count(death_day)
```

Sixty rows came back `NA`: the 59 days before any death total was published, plus the first day that had one, which has no valid predecessor to difference against. `if_else()` gives you a `missing =` argument to label them explicitly:

```{r}
covid %>%
  mutate(
    daily_deaths = deaths - lag(deaths, default = 0),
    death_day = if_else(daily_deaths > 0, "yes", "no", missing = "not reported yet")
  ) %>%
  count(death_day)
```

In `case_when()` you do the same thing with an `is.na()` line placed first, so that missing rows are matched before any comparison has a chance to return `NA`. This keeps "no deaths reported" distinct from "deaths not yet reported"; the two are different findings, and a table that merges them into one category misleads.

## Ordering categories with `factor()`

`count()` is the dplyr way of tabulating a column. Base R's `table()` does the same job in less typing, and it returns a named vector of counts instead of a data frame. Look at the order it puts the labels in:

```{r}
table(covid$severity)
```

The counts come back in alphabetical order — High, Low, Medium — which for an ordered variable carries no meaning. Alphabetical order puts the most severe category first and the middle category last, and any bar chart, legend or summary table built from that column inherits the same ordering.

R sorts alphabetically because `severity` is currently plain text:

```{r}
class(covid$severity)
```

A character vector has no order other than the alphabetical one. `factor()` records that these three strings are the levels of a categorical variable and fixes the order they are held in.

```{r}
covid <- covid %>%
  mutate(severity = factor(severity, levels = c("Low", "Medium", "High")))

table(covid$severity)
```

The counts are 215 Low, 53 Medium and 77 High, and they sum to 345, which is every day of the dataset, so nothing fell through the conditions.

A factor stores small integers plus a lookup table of labels, which is why the ordering sticks:

```{r}
levels(covid$severity)
as.integer(head(covid$severity, 5))
head(covid$severity, 5)
```

Rows 1 to 5 are stored as `1`, displayed as `Low`, and sorted first because `Low` is level one. ggplot reads exactly the same information, so the same fix that repairs `table()` repairs your axes:

```{r}
#| label: fig-severity-order
#| fig-cap: "The same 345 days, counted twice. Only the ordering differs."
p_chr <- ggplot(covid, aes(x = as.character(severity))) +
  geom_bar() +
  labs(title = "Stored as text", x = NULL, y = "Days")

p_fct <- ggplot(covid, aes(x = severity)) +
  geom_bar() +
  labs(title = "Stored as a factor", x = NULL, y = "Days")

p_chr + p_fct
```

The same problem arises for age bands, Likert scales, disease stages, income quintiles, and WHO regions ordered by anything other than the alphabet. Each of them arrives as text, and each needs `factor(levels = ...)` before it tabulates or plots in the intended order.

There is one caution: `factor()` converts any value not listed in `levels` to `NA`, without a warning, including a value that differs only in case:

```{r}
head(factor(as.character(covid$severity), levels = c("low", "Medium", "High")))
table(factor(as.character(covid$severity), levels = c("low", "Medium", "High")), useNA = "ifany")
```

Because the levels say `"low"` where the data has `"Low"`, all 215 Low days become `NA`, without a warning. After creating a factor, run `table(..., useNA = "ifany")` or `count()` once and check that the total still matches the number of rows.

## The dataset, assembled

Putting the pieces together, here is the whole thing in one block: the file as it ships, sorted by date, with daily counts differenced out of the cumulative columns, a severity factor in sensible order, and a `month` column produced by rounding each date down to the first of its month.

```{r}
covid <- read_covid() %>%
  arrange(date) %>%
  mutate(
    daily_cases  = confirmed - lag(confirmed, default = 0),
    daily_deaths = deaths - lag(deaths, default = 0),
    severity = case_when(
      daily_cases <  100 ~ "Low",
      daily_cases <= 300 ~ "Medium",
      daily_cases >  300 ~ "High"
    ),
    severity = factor(severity, levels = c("Low", "Medium", "High")),
    month = floor_date(date, "month")
  )

covid %>%
  select(date, daily_cases, daily_deaths, severity, month) %>%
  slice(85:88)
```

That block is what `covid_daily()` runs. Its definition sits in `_common.R` in the project root, where you can open and read it, so `covid <- covid_daily()` rebuilds this table in one line in a fresh session. It contains nothing beyond what you have now built by hand.

## Exercises

::: {#exr-severity}
## Audit your severity column

Check the `severity` column before relying on it.

1. Confirm the three counts add up to 345 and that no day came back `NA`.
2. For each label, report the smallest and the largest `daily_cases` it contains.
3. Look at the numbers from step 2 either side of each cutoff, and say how far you could move the 100 threshold, in either direction, before a single day changed its label.

For step 3, write down what the answer tells you about how far the labelling depends on the exact value 100.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

`group_by(severity)` followed by a `summarise()` containing `n()`, `min()` and `max()` answers steps 1 and 2 in one call: `group_by()` splits the rows into one pile per label, and `summarise()` computes one row of answers for each pile. If you would rather stay with verbs you have already used, a `filter()` per label and a `range()` on each gets you the same three pairs of numbers with more typing. For the `NA` check, `count()` shows a missing category if one exists.
:::

[Worked solution](../appendices/solutions.qmd#sec-sol-severity)

::: {#exr-move-cutoffs}
## Move the cutoffs

The 100 and 300 thresholds are a convention adopted for this course, and other cutoffs are equally defensible. Rebuild `severity` with cutoffs at 50 and 500 instead, keeping the same boundary conventions (under 50 is Low, 50 up to and including 500 is Medium, over 500 is High).

Report the three counts under the new rule, and compare them with 215 / 53 / 77. Then write two sentences you would be willing to put in a report under each set of cutoffs, describing how severe 2020 was. Both sentences must be true of the same 345 days.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

Copy the `case_when()` block, change the two numbers, and give the new column a different name so you can `count()` both and put the results side by side. Do not forget the `factor()` line, or your new column will sort alphabetically again.
:::

[Worked solution](../appendices/solutions.qmd#sec-sol-move-cutoffs)
