# Reading the epidemic in {#sec-import}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

Everything in this book runs on one file: `sg_covid_data.csv`, 345 rows covering 22 January to 31 December 2020, one row per day. Up to now `read_covid()` has been supplying it. This chapter covers what that function does and what happens at import more generally.

Import is a common source of errors, and most of them are silent. A date column that arrives as text makes the time series plot in alphabetical order; a missing value coded as `-99` that arrives as a number makes the mean wrong; a leading zero disappears from a postcode. None of these raises an error. Most are caught by reading the messages R prints at import time, which is why this book leaves those messages switched on.

## Where the file is {#sec-paths}

R has a working directory: the folder it treats as "here" when you hand it a relative path. It can be set from inside a script, with a line such as `setwd("/Users/swapnil/Documents/GitHub/r-intro")`. That path is specific to one machine, so the script fails for a colleague, after the folder is moved, or on a teaching-lab computer.

`here::here()` avoids the problem. The `here` package looks upward from wherever R happens to be until it finds the top of the project, identified by an `.Rproj` file, a `.git` directory or a `_quarto.yml`, and builds paths from there.

```{r}
#| label: here-root
here()
```

That is this project's root on the machine that rendered this page; on another machine `here()` returns that machine's project root, so paths built from it resolve to the same files. Naming the folders as separate arguments lets it assemble a path with the correct separator for Windows, macOS or Linux.

```{r}
#| label: here-path
here("data", "sg_covid_data.csv")
```

Writing `here("data", "sg_covid_data.csv")` costs a few extra characters and leaves the script working on any machine where the project has been cloned, whatever directory R was started from. Treat this as the permanent habit: no `setwd()` and no absolute paths, in every script.

## `read_csv()` {#sec-read-csv}

readr's `read_csv()` takes a path and returns a tibble.

```{r}
#| label: read
covid <- read_csv(here("data", "sg_covid_data.csv"))
```

The object is named after what it holds: `covid`, not `df` or `data` — the latter masks a base R function. The file itself is read-only: the CSV in `data/` is never edited, and every repair happens in code.

The output above is a message rather than an error, and each line of it is worth reading.

`Rows: 345 Columns: 13`. Check this against what you expect before you do anything else. If you were told the file covers a year of daily data and R reports 345 rows, that is 2020 from 22 January onward, and it adds up. If R had said 344, you would want to know which day went missing and why.

`Delimiter: ","`. readr worked out the separator by looking at the first few lines. Files exported from European locales often use a semicolon, because the comma serves as the decimal mark there; `read_csv2()` handles those.

`chr (1)`, `dbl (11)`, `date (1)`. These are the guessed column types. Three points matter here.

**`date` came in as a real `Date`, not as text.** readr recognises the ISO 8601 form `YYYY-MM-DD` and parses it. Because of that, you can subtract one date from another, sort chronologically, and pass dates directly to `ggplot()`.

**Every count is `dbl`, not integer.** readr's guesser does not distinguish whole numbers from decimals; anything numeric arrives as a double. Nothing goes wrong as a result, but it explains why `confirmed` is `<dbl>` when it counts people.

**One column is `chr`.** That is `administrative_area_level_1`, the constant "Singapore" you met in @sec-tidy.

The two lines at the bottom are hints rather than warnings. `spec()` prints the full specification, one line per column, which is what to use when the summary is truncated with a `...`:

```{r}
#| label: spec
spec(covid)
```

The empty `format = ""` on `date` means readr used its own ISO guesser rather than a format you supplied. The other hint offers `show_col_types = FALSE` to silence the message. That is reasonable once you have read the message and stated the types explicitly, as in @sec-col-types; before then the message is the only report you get on how the file was parsed.

### The dates readr cannot guess

readr guesses dates only for the ISO form. Any other layout is read as character, without a warning, because text is a valid column type and there is nothing for readr to object to.

```{r}
#| label: date-guess
read_csv(I("date,cases\n07/04/2020,106\n08/04/2020,142\n"))
```

The column comes in as `<chr>`. `07/04/2020` is 7 April under day-first conventions and 4 July under month-first ones, and only knowledge of the source can settle which was meant. State the format explicitly, using the same `%d`, `%m`, `%Y` codes that `strptime()` uses:

```{r}
#| label: date-format
read_csv(
  I("date,cases\n07/04/2020,106\n08/04/2020,142\n"),
  col_types = cols(
    date  = col_date(format = "%d/%m/%Y"),
    cases = col_double()
  )
)
```

### Missing values that are not blank

`read_csv()` treats an empty cell and the literal string `NA` as missing, but health data often use other codes. If your file codes missing as `-99`, `.`, `N/A` or `Not reported`, say so, or those values will be read as data:

```{r}
#| label: na-arg
#| eval: false
read_csv(
  here("data", "some_other_file.csv"),
  na = c("", "NA", "N/A", "Not reported")
)
```

With a numeric sentinel such as `-99`, nothing will appear to be wrong. The minimum will be `-99` and the mean will be pulled downward, yet every graph will still render. Checking the minimum and maximum of each numeric column after import will usually reveal a sentinel of this kind.

## `read.csv()` and `read_csv()` {#sec-base-read-csv}

Base R has its own reader, `read.csv()`, spelled with a dot rather than an underscore. It returns a data frame rather than a tibble, it prints no column specification, and it does not parse dates, so the date column arrives as character. The two names differ by one character and are easy to type interchangeably by mistake.

```{r}
#| label: base-read
old <- read.csv(here("data", "sg_covid_data.csv"))
class(old)
class(old$date)
head(old$date, 3)
```

All three differences are visible above: no column specification is printed, `class()` reports a plain `data.frame`, and `date` holds strings that look like dates. The name `old` is itself a weak one, used here only for contrast: it records how the object was read rather than what it holds. Arithmetic on that column fails:

```{r}
#| label: date-arithmetic
#| error: true
old$date - old$date[1]
```

`non-numeric argument to binary operator` reports that the subtraction was attempted on text. The repair is an explicit conversion, `old$date <- as.Date(old$date)`, written before anything touches the column; with `read_csv()` the conversion has already happened at import.

There are two smaller differences. `read.csv()` returns `int` for whole-number columns where `read_csv()` returns `dbl`, and before R 4.0.0 it converted character columns into factors, the historical default described in @sec-tibbles. It is also slower on large files, though at 345 rows the difference is not perceptible.

Both readers load the file. This book uses `read_csv()` because it reports the column types it assigned, and `read.csv()` reports nothing.

## Pinning the types {#sec-col-types}

Guessing is convenient when exploring a new file and less suitable for a script that runs every month. readr guesses from the first 1,000 rows by default, so a column that is numeric for a thousand rows and contains the word `pending` on row 1,500 will be guessed wrong. When the following month's file differs, the guess differs with it, and the code produces different numbers without reporting the change.

The remedy is to state the types. `cols()` lets you name the ones that matter and set a fallback for the rest:

```{r}
#| label: col-types
covid_pinned <- read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols(
    administrative_area_level_1 = col_character(),
    date                        = col_date(format = "%Y-%m-%d"),
    .default                    = col_double()
  )
)
dim(covid_pinned)
```

No specification message appears this time, because there was nothing to guess. The quickest way to write one of these is to read the file once, run `spec()`, and paste the output into the script as the `col_types` argument. The import then either matches the stated specification or reports where it does not.

For example, state that every column is numeric, including the two that are not:

```{r}
#| label: bad-types
bad <- read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols(.default = col_double())
)
```

That produces a warning rather than an error, and the data still loads, with `NA` wherever a value could not be parsed. `problems()` tells you exactly which cells failed:

```{r}
#| label: problems
head(problems(bad))
```

The first failure is at row 2, column 1, where a double was expected and the value `Singapore` was found. Two columns times 345 rows is 690 failures here. When a real import produces a handful of failures rather than hundreds, those few rows are usually the ones worth inspecting: typically a note typed into a numeric field.

When you only need part of a wide file, `cols_only()` reads the named columns and skips the rest, which on a large file saves both time and memory:

```{r}
#| label: cols-only
read_csv(
  here("data", "sg_covid_data.csv"),
  col_types = cols_only(
    date      = col_date(),
    confirmed = col_double(),
    deaths    = col_double()
  )
)
```

The columns come back in the order they appear in the file, not the order you listed them, which matters if any later code indexes columns by position.

### Other formats

Use `read_tsv()` for tab-separated files, `read_delim()` when the separator has to be named explicitly, `readxl::read_excel()` for `.xlsx` files, which takes a `sheet` argument, and the `haven` package for Stata, SPSS and SAS files, which are common in public health. All of them return tibbles, and `haven` retains the variable labels that survey datasets carry.

## Looking at what you loaded {#sec-glimpse}

`glimpse()` is the first thing to run after any import. It transposes the usual view: one line per column, running across the page, so a 13-column table fits on a screen.

```{r}
#| label: glimpse
glimpse(covid)
```

It answers three questions at once: the dimensions confirm the import, the type in angle brackets after each name confirms the parse, and the first few values show the content.

In the values here, seven of the thirteen columns open with `NA`. Two separate things are going on, and the columns are missing for different lengths of time.

```{r}
#| label: na-counts
sum(is.na(covid$deaths))
sum(is.na(covid$parks_percent_change_from_baseline))
```

`deaths` is missing for the first 59 rows. The first value the file carries is 2, on 21 March 2020; the series goes straight to 2 without passing through 1. Everything before that is stored as missing rather than as zero. The two are different claims: zero would state that nobody died, while `NA` states that the source does not report a figure. The distinction costs two deaths as soon as the column is differenced to recover daily deaths, because the first recorded value of 2 has `NA` behind it and there is nothing to subtract from. The daily figures then sum to 27 against a recorded total of 29.

The six mobility columns are missing for the first 24 rows, up to and including 14 February. Google's Community Mobility Reports express each day as a percentage change from a baseline computed over 3 January to 6 February 2020, so the series cannot start until that baseline period has finished. Neither gap is a fault in the file; in both cases the source has no value to report. `glimpse()` makes both visible immediately after loading.

Three further functions are worth knowing. `head()` and `tail()` show the first and last rows, and `tail()` is the one that catches a summary row left at the bottom of a spreadsheet. `summary()` gives a five-number summary of every numeric column plus a count of `NA`s. `View()` opens the spreadsheet viewer in RStudio, which is convenient for a small file and unhelpful on a large one; keep it at the console rather than in a script, because it does nothing in a rendered document.

```{r}
#| label: view
#| eval: false
View(covid)
```

## The helper this book uses {#sec-read-covid}

`read_covid()` does nothing beyond what this chapter has covered. It calls `read_csv()` on this file via `here()`, with the column-types message switched off because the specification is set out above:

```{r}
#| label: helper
#| eval: false
read_covid <- function() {
  readr::read_csv(here::here("data", "sg_covid_data.csv"), show_col_types = FALSE)
}
```

The name follows the same convention as object names: a function is named for what it does, and @sec-functions covers how to write your own. It does no cleaning or repair: counts are still cumulative and `NA`s are still `NA`s. Where a later chapter corrects something, it does so in code you can read.

## A first plot {#sec-first-plot}

`ggplot()` is covered in a later chapter. The code below is included here so that the file can be seen as a whole before any of it is transformed.

```{r}
#| label: reward-plot
covid <- read_covid()

ggplot(covid, aes(x = date, y = confirmed)) +
  annotate(
    "rect",
    xmin = as.Date("2020-04-07"), xmax = as.Date("2020-06-01"),
    ymin = -Inf, ymax = Inf,
    fill = nus_colours[["grey"]], alpha = 0.4
  ) +
  geom_line(linewidth = 0.8) +
  annotate(
    "text",
    x = as.Date("2020-05-04"), y = 50000,
    label = "circuit breaker", size = 3.2, colour = nus_colours[["muted"]]
  ) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b") +
  labs(
    title = "Singapore's 2020 epidemic, exactly as the file records it",
    subtitle = "Cumulative confirmed cases. Every point is a running total, not a daily count.",
    x = NULL,
    y = "Confirmed cases",
    caption = "Source: sg_covid_data.csv, 22 January to 31 December 2020."
  )
```

By the end of December the file records fifty-eight thousand confirmed cases, and the curve turns sharply upward in April before flattening through the second half of the year. The shaded band is Singapore's circuit breaker, 7 April to 1 June 2020.

The curve records the file accurately, but a cumulative total can only increase, so it shows neither when transmission accelerated nor when it stopped. The quantity that does is the daily difference, each day's total minus the previous day's, held in a column of its own. On 20 April that column holds 1,426 new cases; the cumulative line for the same day holds 8,014, which describes January to April rather than 20 April.

## Exercise

::: {#exr-first-look}
## Land the file yourself

Start a fresh R session, load the tidyverse and `here`, and do this without using `read_covid()`.

1. Read `data/sg_covid_data.csv` with `read_csv()` and a path built by `here()`. Write down, in words, the type readr assigned to `date`, to `confirmed` and to `administrative_area_level_1`, and where in the message you found each one.
2. Read it a second time with the three types above stated explicitly and everything else left as double, and confirm the specification message no longer appears.
3. Run `glimpse()` and count the columns whose first value is `NA`. Then find the number of missing values in each of those columns, and say whether they all have the same number.
4. Read the same file with `read.csv()` instead. Find the one difference in the `date` column that would have forced you to write extra code, and demonstrate it failing.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

For part 2, `cols()` takes named columns plus a `.default` for everything you did not name. For part 3, `is.na()` returns a logical vector and `sum()` counts the `TRUE`s. For part 4, `class()` on the column is the fastest tell, and the failure you want is any arithmetic on a date.
:::

[Worked solution](../appendices/solutions.qmd#sec-sol-first-look)
