# Where to go next {#sec-next-steps}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

```{r}
#| label: data
#| include: false
covid <- covid_daily()
```

A week of R gets you a working vocabulary. You can read a file, check it, reshape it, summarise
it by group, plot it, and write a function when the same six lines show up for the third time.
What you do not yet have is the accumulated set of small recoveries that make an error message an
inconvenience rather than a stopping point. Those come from hours at the keyboard, and no chapter
can supply them. This last chapter describes where those hours are best spent: how the modelling
weeks build on what you have learned, what to read next, what to do when code fails, and how to
ask a question that gets answered.

## What the rest of the course does with all this

The modelling weeks turn the direction of travel around. Everything so far started with data and
went looking for a description. A compartmental model starts with a description, some
assumptions about how people move between states, and produces data you can compare with the
real thing.

The mechanics are less exotic than they sound. You divide a population into compartments,
susceptible, infectious, recovered in the simplest case. You write down the rate at which people
leave each compartment and enter the next. Then you hand those rates to a solver, which steps
them forward in small increments and hands back a trajectory.

In R the solver is usually [deSolve](https://cran.r-project.org/package=deSolve), and its main
function is `ode()`. The shape of the code is this.

```{r}
#| label: sir-sketch
#| eval: false
sir <- function(t, state, parameters) {
  with(as.list(c(state, parameters)), {
    dS <- -beta * S * I / N
    dI <-  beta * S * I / N - gamma * I
    dR <-  gamma * I
    list(c(dS, dI, dR))
  })
}

state      <- c(S = 99999, I = 1, R = 0)
parameters <- c(beta = 0.4, gamma = 0.2, N = 1e5)
times      <- seq(0, 180, by = 1)

out <- ode(y = state, times = times, func = sir, parms = parameters)
```

Read that against what you already know and almost all of it is familiar. `sir` is a function
with three arguments whose value is whatever the last expression evaluates to, which is the
functions chapter. `state` and `parameters` are named numeric vectors, and `as.list()` and
`with()` are there so you can write `beta` instead of `parameters[["beta"]]`, which is the
foundations chapter. The return value is a list containing one vector, because `ode()` insists
on a list, and knowing the difference between `c(dS, dI, dR)` and `list(c(dS, dI, dR))` is why
that chapter spent time on lists.

What `ode()` gives back is a matrix, not a tibble. Your first move will be `as_tibble()`, then
`pivot_longer()` so you can draw three lines on one plot, then the ggplot2 grammar from the
visualising chapter. Two evenings of this course will be spent on models and about ten minutes
of each of them on plotting, because you already did the plotting.

Two things will catch you, so they are worth naming now.

The `time` column that comes out of `ode()` counts steps from zero. It is not a date. If you
want to line a trajectory up against the `daily_cases` column you built, you have to decide what
day zero means and add it: `start_date + out$time`. Model time and calendar time are different
quantities, and mixing them silently is a common way to produce a plot that looks fine and is
wrong by a fortnight.

The second is that you will want to run the model more than once. Sweeping over ten values of
beta, or a hundred, is the iteration chapter's problem exactly: write the model run as a function
of the parameter you are varying, then map over the values and bind the results, rather than
copying the call ten times with a number changed.

Once your models get big enough that speed matters, or you are running thousands of stochastic
replicates, R's own loop overhead starts to hurt. [odin](https://cran.r-project.org/package=odin)
is the usual answer: you write the equations in a small R-like language and it generates and
compiles C for you. The equations look much like the ones above. The habits of writing them
carefully, naming parameters properly and checking a trajectory before believing it are the same
habits.

Fitting a model to Singapore's 2020 curve is where the two halves of the course meet, and it is
also where the caveats from the capstone come back. You would be fitting to reported cases, which
are a function of testing as well as transmission, over a period when the national count was
dominated by the dormitory outbreak. A model fitted to that series without a term for either will
absorb both into its transmission parameters.

## Five things worth reading

The list is deliberately short, and each entry states what the book covers and when it is worth
reading.

[**R for Data Science**, 2nd edition](https://r4ds.hadley.nz/), by Wickham, Çetinkaya-Rundel and
Grolemund. Free online. This is the book this course is a one-week introduction to. The chapters on
data transformation and visualisation cover the same ground with more room, and the chapters on
joins, factors, strings and dates cover the topics this book leaves out. If you read one item from
this list, make it this one, and work through it at the keyboard.

[**The Epidemiologist R Handbook**](https://epirhandbook.com/en/), by Neale Batra and a large
group of contributors. Free online. This is a recipe book rather than a textbook, written by
field epidemiologists for field epidemiologists, and it is the fastest route from "I need an epi
curve with a moving average and a case fatality ratio by age group" to working code. Its examples
use outbreak data. Use it when you know what you want but not which function produces it.

[**Happy Git and GitHub for the useR**](https://happygitwithr.com/), by Jenny Bryan. Free online.
Version control repays the effort of learning it quickly, and this is the book that gets R users
past the initial difficulty of git. Read the first eight or so chapters and get one project under
version control; that is enough for now. You do not need to understand rebasing. The everyday Git
cycle is set out in @sec-good-practice.

[**The tidyverse style guide**](https://style.tidyverse.org/), by Hadley Wickham. It takes an
afternoon to read and much of it can be skimmed. It is where the conventions this book uses come
from: spaces around `<-`, lowercase names with underscores, one dplyr verb per line. Consistency
earns its keep because it lets you read your own code from three months ago without decoding it
first.

[**Advanced R**, 2nd edition](https://adv-r.hadley.nz/), by Hadley Wickham. Free online. This one
is not for now; bookmark it and return to it when you find yourself asking why something behaves
as it does rather than how to make it work. The chapters on environments and on functionals are
the ones that explain the odd corners of R that this book has been telling you to accept.

Two kinds of book are deliberately absent from this list: there is no reference manual, because
R's own help pages are the reference manual and `?filter` is faster than any book; and there is
no book on statistics in R, because the statistics course covers the statistics and the R
involved in it is the straightforward part.

## When something breaks

When something breaks, work through the steps in this section in the order given.

### Read the whole error message

R's error messages have improved a great deal, and the tidyverse ones in particular are built to
be read from top to bottom rather than skimmed for the first line. The code below produces one
deliberately.

```{r}
#| label: broken
#| error: true
month_summary <- function(data, month_start) {
  data %>%
    filter(month == month_start) %>%
    summarise(cases = sum(daily_cases), peak = max(daily_cases))
}

month_summary(covid, "2020-08")
```

The message has four lines, each carrying a separate piece of information. The first says which
verb failed: `filter()`,
not `summarise()`, so the pipeline died before it got to the summary. The reference to
`dplyr/R/summarise.R` is dplyr telling you where in its own source the call was recorded, and it
is not about your file. The second line names the argument that failed, `month == month_start`.
The third says the actual failure happened inside `charToDate()`, which is a base R function for
turning text into dates and not a function you called. The fourth is the complaint itself.

Put those together and the account is complete. Comparing a `Date` column with a character string
makes R try to convert the string to a date, and `"2020-08"` is not a date because R has no way
of telling whether you meant the first of the month or the whole of it, and it refuses to guess.

```{r}
#| label: fixed
month_summary(covid, as.Date("2020-08-01"))
```

When you read an error, work out which line names *your* code and which lines name the machinery
underneath it. The fix is almost always at the boundary between them.

If you are working in RStudio you will also see `Run rlang::last_trace() to see where the error
occurred.`, and it is worth running. It prints the full chain of calls with your own function at
the top and the failing one at the bottom, and for a pipeline buried inside a function it is
often the quickest way to find which of your six lines broke.

### `traceback()`, for errors that do not say where they occurred

Base R errors are terser, as in this example from a function calling another function.

``` r
> daily_mean <- function(x, i, k) mean(x[(i - k):(i + k)])
> smooth_at <- function(data, i) daily_mean(data$daily_cases, i, 3)
> smooth_at(covid, "100")
Error in i - k : non-numeric argument to binary operator
```

That message says what went wrong but not where; `traceback()`, run immediately
afterwards, identifies where.

``` r
> traceback()
3: mean(x[(i - k):(i + k)]) at #1
2: daily_mean(data$daily_cases, i, 3) at #1
1: smooth_at(covid, "100")
```

Read it from the bottom. Line 1 is the call you typed. Line 2 is what it called. Line 3 is where
it died. So `smooth_at()` passed the string `"100"` straight through to `daily_mean()`, which
tried to subtract 3 from it. The `at #1` is the line number in the source; when your functions
live in a file rather than being typed at the console, you get something like `at report.R#12`,
which is the number you want.

RStudio puts a "Show Traceback" button next to errors in the console, which does the same thing
in one click.

### `browser()`, for inspecting a function as it runs

When the traceback tells you which function died but not why, put `browser()` on the line before
the trouble and run the call again. R stops there and hands you a prompt with the function's own
variables in scope.

```{r}
#| label: browser-demo
#| eval: false
month_summary <- function(data, month_start) {
  browser()
  data %>%
    filter(month == month_start) %>%
    summarise(cases = sum(daily_cases))
}

month_summary(covid, "2020-08")
```

``` r
Called from: month_summary(covid, "2020-08")
Browse[1]> month_start
[1] "2020-08"
Browse[1]> class(month_start)
[1] "character"
Browse[1]> class(data$month)
[1] "Date"
Browse[1]> Q
```

Three questions identify the bug: you are comparing a character against a Date. Type `Q` to
quit the browser, `n` to run the next line, `c` to let it finish. Then remove the `browser()`
call. Left in place, it produces a script that keeps stopping at the same line for no visible
reason.

`debugonce(month_summary)` does the same thing without editing the function, which is preferable
when the function belongs to a package.

### Failures that produce no error

An error is the easier case, because R stops and reports what went wrong. The failures that
cause more trouble are the ones that return a number, because nothing marks them as wrong.

```{r}
#| label: silent
covid$date[which.max(covid$daily_case)]
```

The cause is one missing letter in the column name. R gives you a warning and a zero-length date rather than
an error, because `$` on a tibble returns `NULL` for a column that is not there and
`which.max()` of a zero-length input returns a zero-length result. Inside a longer pipeline that
warning scrolls past and you publish a report with a blank where the peak date should be.

The same problem arises in this dataset, where two reasonable-looking calls give different
answers.

```{r}
#| label: deaths-gap
max(covid$deaths, na.rm = TRUE)
sum(covid$daily_deaths, na.rm = TRUE)
```

Singapore recorded 29 COVID-19 deaths in 2020. Differencing the cumulative column recovers 27 of
them, because `deaths` has 59 leading missing values and the first number that appears is already
2, on 21 March. The two missing deaths are hidden inside that first observation. Neither line
above produces an error or a warning. One of them is the wrong answer to the question "how many
people died", and which one depends on how that question is meant.

What catches these is arithmetic against something you already know. Check the total against a
number you have from another source. Check that the number of rows matches the number of days.
Check that the maximum lands on the date you expected. Each of those is a line of code, and each
catches mistakes that reading the code will not.

## How to ask a question that gets answered

You will get stuck in a way that fifteen minutes of your own effort will not fix. When that
happens, how the question is put makes a large difference to how quickly it gets answered.

A question that gets answered has four parts: what you were trying to do, stated in terms of the
result you wanted; the code you ran, as text; what happened, with the error message copied
exactly; and what you have already tried, so the answers you get start from where you stopped.

The exact error text matters more than anything else in that list. Post the code and the error
text rather than describing them. "It says something about a non-numeric argument" cannot be
searched for. `non-numeric argument to binary operator` can, and the search will usually find
other people who have had the same problem. Before you ask anyone, put the message in a search
engine with the word `r` and the name of the package. That resolves a large share of problems on
its own.

Reduce the example to the smallest code that still fails. If your pipeline is forty lines, cut it
in half and see whether the error survives, then repeat. Most bugs reduce to three or four lines,
and the process of cutting often finds the answer, so it is worth doing even when you intend to
ask anyway. If the problem needs data, do not attach a 30 MB CSV: `dput(head(covid, 5))` prints a
few rows as code that anyone can paste into their own session.

The [reprex](https://cran.r-project.org/package=reprex) package automates the tidy version of
this. You copy your minimal example to the clipboard, run `reprex::reprex()`, and it runs the
code in a clean session and puts a formatted block on your clipboard with the code and its real
output interleaved, ready to paste.

```{r}
#| label: reprex-demo
#| eval: false
reprex::reprex({
  library(dplyr)
  x <- c(1, 2, NA)
  mean(x)
})
```

The clean session is what makes this worth doing. It catches the case where your code only works
because of something you loaded an hour ago and forgot about, which accounts for a large share of
problems that look like package bugs.

Include your R version and package versions when the problem may be a version difference.
`sessionInfo()` prints them.

Classmates are the first place to ask. They are working on the same data, the same package
versions and the same week, so someone is likely to have met the same problem already. After
that, the [Posit Community](https://forum.posit.co/) forum is patient with beginners, and Stack
Overflow is faster but less forgiving, so read a few answered R questions there before posting
your first one. Report an issue on a package's GitHub only once you can show the maintainer a
reproducible example and you are reasonably confident the package is at fault rather than your
own code.

Post code as text rather than as a photograph of the screen. Code cannot be copied out of a
photograph, so answering your question would mean retyping it first.

When you work something out that took two hours, record the fix in a comment in the script, in
the words you would have needed at the start. You will meet the same problem again, and the
comment is what makes the second encounter a short one.
