# Rectangles: data frames and tibbles {#sec-rectangles}

```{r}
#| label: setup
#| include: false
source("_common.R")
```

A data frame holds several variables of different types in one object, with the same number of rows in each. That constraint is what allows a dataset to be filtered, sorted and summarised without keeping track by hand of which value belongs with which. This chapter covers what a data frame is made of, what a tibble changes, and what tidy data means.

```{r}
#| label: load
covid <- read_covid()
```

## Three vectors and a problem {#sec-loose-vectors}

Take three days of the Singapore record, in the week the circuit breaker began, and store each variable in its own vector.

```{r}
#| label: loose
day        <- as.Date(c("2020-04-06", "2020-04-07", "2020-04-08"))
confirmed  <- c(1375, 1481, 1623)
stringency <- c(45.37, 62.04, 76.85)
```

The values themselves are all present, but the correspondence between the three vectors is not recorded anywhere: that the second element of each refers to the same day is an assumption you are holding, not something R knows or enforces.

Sorting one of them shows the consequence. Sort the stringency values from highest to lowest, which is a reasonable thing to want:

```{r}
#| label: sort-breaks
sort(stringency, decreasing = TRUE)
day
```

`stringency` has been reordered and `day` has not, so if that sorted vector is assigned back over the old one, the 7 April stringency score now sits alongside 6 April, and nothing in the session records that the two vectors have come apart. There is no error and no warning, and a plot drawn from them still looks plausible.

With 345 rows and 13 variables, maintaining that correspondence by hand is not workable, which is what data frames are for.

## A data frame is a list of columns {#sec-list-of-columns}

`data.frame()` binds vectors together so that they can only be manipulated as a unit.

```{r}
#| label: build-df
three_days <- data.frame(day, confirmed, stringency)
three_days
```

Now sorting means sorting *rows*, and the rows carry their contents with them:

```{r}
#| label: sort-rows
three_days[order(three_days$stringency, decreasing = TRUE), ]
```

The row numbers on the left have gone 3, 2, 1. Each row moved as a piece.

Underneath, a data frame is a list of vectors, which becomes visible if you strip the class off it:

```{r}
#| label: unclass
unclass(three_days)
```

The output is a named list of three vectors plus an attribute recording the row names, and that is all a data frame contains. Everything else, the printing, `nrow()`, the two-dimensional subsetting, is behaviour attached to that list by its class. R will confirm the underlying structure if asked directly:

```{r}
#| label: is-list
is.list(three_days)
length(three_days)
lengths(three_days)
```

`length()` on a data frame counts the elements of the underlying list, which is the number of columns. `lengths()`, with the s, gives the length of each element, which is the number of rows repeated once per column. `nrow()` and `ncol()` are friendlier names for the same two facts.

Two properties follow directly from the list-of-vectors implementation.

**Columns may be of different types.** A list can hold anything, so one element can be a `Date` vector, the next a `double` vector, the next a `character` vector. Each column is internally uniform, which is the vector rule from @sec-types, but columns need not agree with each other. That is what makes a data frame a suitable shape for a dataset, where dates, counts and labels sit side by side. A matrix cannot do this; every cell of a matrix has one type.

```{r}
#| label: col-types
sapply(three_days, class)
```

**Rows must line up.** The list is not free to hold vectors of any length, because the rectangle would stop being a rectangle. Give `data.frame()` mismatched lengths and it refuses:

```{r}
#| label: mismatch
#| error: true
data.frame(day = 1:3, phase = c("pre", "circuit breaker"))
```

`arguments imply differing number of rows: 3, 2` is a useful error: it arrives at the moment the mistake is made rather than several steps later.

Every operation on a rectangle, filtering rows, creating columns, grouping and summarising, relies on row *i* of every column referring to the same observation. `filter()` can drop 200 rows without any separate check that the variables stayed aligned, because misalignment is not representable.

### Recycling in `data.frame()`

There is one exception to that constraint, and it is a long-standing one. If the short vector's length divides the long one exactly, `data.frame()` recycles instead of raising an error:

```{r}
#| label: recycle-df
data.frame(day = 1:6, phase = c("pre", "circuit breaker"))
```

The result is six rows with alternating labels, produced without a warning. This is the recycling rule from @sec-recycling applied to column construction. It is convenient when intended and silent when not, so a mistyped or mis-subsetted vector can produce a table that looks complete.

## Tibbles {#sec-tibbles}

A tibble is a data frame with several of its older behaviours changed. `read_csv()` returns one, so this is what `covid` has been all along:

```{r}
#| label: class-covid
class(covid)
```

Reading the class vector from the right, the object *is* a `data.frame`, so anything that accepts a data frame accepts it. `tbl` and `tbl_df` are the tibble classes, which change how it prints and how it is subsetted. `spec_tbl_df` is readr's marker, carrying the column specification it worked out while reading the file: which column it took to be a date, which it took to be numbers, which it left as text.

The differences that matter in practice are the following.

### Printing reports types and dimensions

```{r}
#| label: print-tibble
covid
```

The printout shows ten rows, then a count of what was left out, then the names and types of the columns too wide to show. The `<chr>`, `<dbl>` and `<date>` labels under the column names are worth attention: a column that is character when you assumed it was numeric is a frequent source of confusion, and the tibble reports the type before any analysis is written.

An old-style data frame does not do this. Print the same 345 rows as a `data.frame` and R sends all of them to the console without type labels, so the header scrolls off the screen. It is worth doing once to see what it looks like, then using `head()`:

```{r}
#| label: print-df
head(as.data.frame(covid)[, 1:5])
```

### No accidental factors

Before R 4.0.0, released in April 2020, `data.frame()` and `read.csv()` converted every character column into a factor by default, because the `stringsAsFactors` argument defaulted to `TRUE`. A factor stores its values as integer codes against a table of levels, which is appropriate for a categorical variable in a model and not for a column of free text or identifiers.

The behaviour is still available on request, and it has one consequence that is easy to overlook:

```{r}
#| label: factors
phase <- data.frame(label = c("pre", "circuit breaker"), stringsAsFactors = TRUE)
phase$label
```

`Levels: circuit breaker pre` is alphabetical, so "circuit breaker" sorts before "pre" and any plot will show them in that order. R 4.0.0 changed the default to `FALSE`, one of the few backwards-incompatible changes the language has made. Tibbles never performed the conversion.

Factors are useful when requested deliberately. A severity label that should read low, medium, high rather than alphabetically is a case for `factor()`, because you name the levels in the order you want and everything downstream uses that order. The difference is that the conversion is then explicit in the code.

### Subsetting stays a data frame

Single-bracket subsetting of one column behaves differently in the two classes.

::: {.callout-important}
## Predict before you run

`old` below is the same data as `covid`, converted to an old-style data frame. Both of the next two expressions ask for one column with a single pair of square brackets. Decide, before running them, whether each returns a rectangle of 345 rows and 1 column, or a bare vector of 345 numbers.
:::

```{r}
#| label: drop-setup
old <- as.data.frame(covid)
class(old[, "confirmed"])
class(covid[, "confirmed"])
```

The data frame dropped a dimension and returned a vector, whereas the tibble returned a one-column table.

```{r}
#| label: drop-show
covid[, "confirmed"]
```

The old behaviour comes from `[.data.frame` having a `drop` argument that defaults to `TRUE`: ask for a single column and R returns its contents rather than a one-column table. That is convenient interactively and a problem inside a pipeline, because a function two steps downstream that expected a data frame receives a numeric vector and fails with a message about something else entirely. Tibbles set `drop = FALSE`, so `[` always returns a tibble.

When you do want the bare vector, say so with `$` or double brackets, both of which mean "give me the contents of this element" and work identically on both kinds of rectangle:

```{r}
#| label: extract
class(covid$confirmed)
class(covid[["confirmed"]])
```

### No silent partial matching

`$` on an old-style data frame will complete a partial column name for you:

```{r}
#| label: partial-df
head(old$conf, 5)
```

There is no column called `conf`. R found `confirmed`, took it as the intended match, and returned it without comment. If the file also had a `confirmed_cases_dormitory` column the match would be ambiguous and `$` would return `NULL`, so the same line would behave differently on a file that differs only in having an extra column.

Tibbles warn:

```{r}
#| label: partial-tbl
head(covid$conf, 5)
```

The result is a warning saying the column is unknown or uninitialised, followed by `NULL`. The same warning appears for an ordinary typo, so `covid$confrimed` is reported rather than silently returning `NULL` into a calculation.

### Choosing between them

Use tibbles for anything you write. When an older package rejects one, and a few statistical packages still do, convert at the point of the call:

```{r}
#| label: as-data-frame
class(as.data.frame(covid))
```

`as_tibble()` converts back. The two are interconvertible, so the choice is local to the call rather than a commitment for the whole project.

## Tidy data {#sec-tidy}

Tidy data is a rule about what goes into which part of the rectangle, and it has three clauses:

1. Each variable is a column.
2. Each observation is a row.
3. Each value is a cell.

The three clauses are easy to state, but real spreadsheets violate them routinely: twelve columns named January to December, a column holding both a measurement and its unit, or two rows per patient because before and after were recorded side by side.

Tidyverse functions assume tidy input. Once the data are in that shape, `filter()`, `mutate()`, `group_by()` and `ggplot()` all work without further reshaping.

Applying the definition to this file means first establishing what one observation is:

```{r}
#| label: unit
nrow(covid)
length(unique(covid$date))
length(unique(covid$administrative_area_level_1))
```

The output shows 345 rows, 345 distinct dates and one country, so the unit of observation is a country-day; since the country never varies, in practice it is a day. Each column is a variable measured on that day: a cumulative case count, a cumulative death count, a policy index, six mobility series. There is one row per observation, one column per variable and one value per cell, so this file arrives tidy, which is convenient for the book; data you are handed elsewhere often will not be.

### A column that says nothing

`administrative_area_level_1` deserves a second look:

```{r}
#| label: constant-col
table(covid$administrative_area_level_1)
```

The column has one distinct value across all 345 rows. A column that never varies cannot explain, predict, group or distinguish anything. It is present because this extract was taken from a multi-country archive in which the column did vary, and selecting a single country left the label behind.

Counting distinct values in every column is a habit worth forming on any new dataset, because both ends of the result tell you something. A column with one distinct value carries no information. A column with as many distinct values as there are rows is an identifier rather than a measurement, which is what `date` is here. Anything in between is a variable you can actually work with.

```{r}
#| label: distinct-counts
sapply(covid, function(column) length(unique(column)))
```

The `function(column)` argument is an anonymous function, defined at the point of use rather than given a name; writing your own functions is covered in @sec-functions.

`stringency_index` takes 13 distinct values over the year, which reflects how it is constructed: it is a score assembled from a handful of policy indicators, so it changes only when one of those indicators changes. The series therefore moves in steps, and each step marks a discrete policy change.

### When the file is not tidy

Untidy files are common. The usual shapes are a header row of dates or months, one column containing two variables combined, and observations split across several rows. `tidyr` provides `pivot_longer()` and `pivot_wider()` for reshaping between wide and long layouts. This book does not devote a chapter to reshaping, because the Singapore file does not require it, and the operations are easier to follow once the material here is familiar.

## Exercise

::: {#exr-tibble-anatomy}
## Take a tibble apart

Using `covid` as loaded at the top of this chapter:

1. Show, with code rather than assertion, that `covid` is a list, how many elements that list has, and that every element has the same length. Say in one sentence which of those two numbers is the number of rows and which is the number of columns.
2. Find out the type of every column without printing the whole tibble.
3. Build a small tibble by hand recording three dates in Singapore's 2020 policy timeline, 7 April, 2 June and 19 June, alongside a short label for each. Give the columns snake_case names that say what each column holds. Then explain in two sentences why holding those as a tibble is safer than holding the dates and the labels as two separate vectors.
4. Convert `covid` to an old-style data frame and find one behaviour, other than printing, that differs from the tibble. Show it.
:::

::: {.callout-note collapse="true"}
## Stuck? Open for a hint (not the answer)

For part 1, three functions from this chapter do the whole job: one asks whether it is a list, one counts the elements, one measures each element. For part 2, `sapply()` applies a function to every column and collects the answers into one named vector.
:::

[Worked solution](../appendices/solutions.qmd#sec-sol-tibble-anatomy)
